---
title: "Regression Analysis - STAT 482 - Probem Set 10"
author: "Alex Towell (atowell@siue.edu)"
header-includes:
- \usepackage{amsmath}
- \usepackage{mathtools}
- \usepackage{amsthm}
- \usepackage{multirow}
- \usepackage{booktabs}
- \usepackage{minted}
- \usepackage{color}
- \usepackage{xcolor}
- \usepackage{tcolorbox}
- \usepackage{enumerate}
output:
  pdf_document:
    toc: no
    latex_engine: xelatex
    df_print: kable
editor_options:
  markdown:
    wrap: 80
---

```{=tex}
\newcommand{\sos}[1]{\mathrm{SS_{#1}}}
\newcommand{\ms}[1]{\mathrm{MS_{#1}}}
\newcommand{\sd}{\operatorname{sd}}
\newcommand{\var}{\operatorname{var}}
\newcommand{\expect}{\operatorname{E}}
\newcommand{\corr}{\operatorname{cor}}
\newcommand{\cov}{\operatorname{cov}}
\newcommand{\se}{\operatorname{se}}
\newcommand{\eval}[2]{\left. #1 \right\vert_{#2}}
\newcommand{\degf}[1]{\mathrm{df_{#1}}}
\newcommand{\entropy}{\operatorname{H}}
\newcommand{\param}[1]{\textcolor{blue}{\texttt{#1}}}
\newcommand{\ci}{\operatorname{CI}}
```

```{r, include=FALSE}
options(tinytex.engine_args = '-shell-escape')
```

> A company wishes to study the effects of three different types of promotions
> on sales of its crackers.
> For each store in the sample, the sales for the promotion period ($y$) and the
> sales for the preceeding period ($x$) are observed.
> The data is available from Table 22.1 on the course website.
> Response variable $y$ is listed in the first column, continuous input $x$ is
> listed in the second column, categorical input promotion type (1,2,3) is
> listed in the third column.
> The fourth column, the observation number, can be ignored.

## Preliminary
We load the data in R with:
```{r}
data = read.table('CH22TA01.txt')[,1:3]
colnames(data) = c("sales","pre.sales","type")
data$type = as.factor(data$type)
head(data)
```

## Problem 1
> Define indicator variables $I_1$ and $I_2$ using promotion type 3 as the
> baseline level.

We let
$$
  I_1 =
  \begin{cases}
    1, & \text{if} \;\param{type} = 1,\\
    0, & \text{otherwise}
  \end{cases}
$$
and
$$
  I_2 =
  \begin{cases}
    1, & \text{if} \;\param{type} = 2,\\
    0, & \text{otherwise}.
  \end{cases}
$$

We do this transformation in R with:
```{r}
contrasts(data$type) = contr.treatment(3,base=3)
contrasts(data$type)
```

## Problem 2
> Write an additive model for response $y$ using continuous input variable $x$ and
> indicator variables $I_1$, $I_2$.

The model of the mean $\param{sales}$ given $\param{pre-sales}$ and $\param{type}$
is given by
$$
  E(\param{sales}) = \beta_0 + \beta_1 \param{pre-sales} + \beta_2 I_1 + \beta_3 I_2.
$$

### Comments
The model of the data generating process given the predictors is given by
$$
  \param{sales}_i = \beta_0 + \beta_1 \param{pre-sales} + \beta_2 I_1 + \beta_3 I_2 + \epsilon_i,
$$
where $\epsilon_i$ is zero mean i.i.d. normal.

## Problem 3
> Write a regression function for each of the promotion types.

$$
  E(\param{sales}) = 
  \begin{cases}
    (\beta_0 + \beta_2) + \beta_1 \param{pre-sales} & \text{if} \;\param{type} = 1,\\
    (\beta_0 + \beta_3) + \beta_1 \param{pre-sales} & \text{if} \;\param{type} = 2,\\
    \beta_0 + \beta_1 \param{pre-sales}         & \text{if} \;\param{type} = 3.
  \end{cases}
$$

## Problem 4
> Define predictor effect parameters as partial differences in means.

Letting $y$ denote $\param{sales}$, $x$ denote the continuous predictor
$\param{pre-sales}$, and $t_j$ denote $\param{type}=j$.

Then,
\begin{align*}
  \beta_1 &= \frac{\partial E(y)}{\partial x}\\
  \beta_2 &= E(y|x,t_1) - E(y|x,t_3)\\
  \beta_3 &= E(y|x,t_2) - E(y|x,t_3)\\
  \beta_2-\beta_3 &= E(y|x,t_1) - E(y|x,t_2).
\end{align*}

## Problem 5
> Provide an interpretation for each effect parameter, stated in the context of
> the problem.

$\beta_1$ is the difference in mean \param{sales} from a $1$ unit
increase in \param{pre-sales}, with the \param{type} held constant.

$\beta_2$ is the difference in mean \param{sales} between \param{type} $1$
and \param{type} $3$, with \param{pre-sales} held constant.

$\beta_3$ is the difference in mean \param{sales} between \param{type} $2$
and \param{type} $3$, with \param{pre-sales} held constant.

$\beta_2-\beta_3$ is the difference in mean \param{sales} between \param{type} $1$
and \param{type} $2$, with \param{pre-sales} held constant.


## Problem 6
> Compute interval estimates for each of the effect parameters.

To derive $\ci(\beta_2 - \beta_3)$, we replicate a bit of distribution theory.
Let $a = (0\;0\;1\;-1)'$ and $b = (b_0\;b_1\;b_2\;b_3)$, then
$$
  a' b = b_2 - b_3,
$$
for which
$$
  E(a' b) = a' E(b) = (0\;0\;1\;-1) (\beta_0\;\beta_1\;\beta_2\;\beta_3)' = \beta_2 - \beta_3
$$
and
$$
  \var(a' b) = a' \cov(b) a = \sigma^2 a' (X'X)^{-1} a
$$
where
$$
  X =
  \begin{Bmatrix}
    1 & x_1 & I_{1 1} & I_{1 2}\\
    1 & x_2 & I_{2 1} & I_{2 2}\\
    \vdots & \vdots & \vdots\\
    1 & x_n & I_{n 1} & I_{n 2}.
  \end{Bmatrix}
$$

Note that we do not know $\sigma^2$, so we estimate it with $\ms{E}$.

We compute the estimates of the coefficients in the additive model with:
```{r}
library(matlib)
add.mod = lm(sales ~ pre.sales+type,data=data)

b.hat = coef(add.mod)
dfe = nrow(model.matrix(add.mod)) - ncol(model.matrix(add.mod))
V = vcov(add.mod)
a = c(0,0,1,-1)

b.hat.12 = a %*% b.hat
se.12 = sqrt(a %*% V %*% a)

#t.stat.12 = b.hat.12 / se.12
#p.value.12 = 2*(1-pt(abs(t.stat.12),dfe))
#print(c(t.stat.12,p.value.12))

b.hat.12.lower = b.hat.12 - qt(.975,dfe) * se.12
b.hat.12.upper = b.hat.12 + qt(.975,dfe) * se.12

print(c(b.hat.12.lower,b.hat.12.upper))
confint(add.mod)
```

We see that
\begin{align*}
  \ci(\beta_1) = [.673,1.124],\\
  \ci(\beta_2) = [10.323,15.630],\\
  \ci(\beta_3) = [5.285,10.518],\\
  \ci(\beta_2-\beta_3) = [2.370, 7.780].
\end{align*}

## Problem 7
> Create a scatterplot of the data with the estimated regression lines.

```{r}
intercept.3 = b.hat[1]
intercept.1 = b.hat[1]+b.hat[3]
intercept.2 = b.hat[1]+b.hat[4]
slope = b.hat[2]

cat(slope, intercept.1, intercept.2, intercept.3)

attach(data)
plot(pre.sales[type == 1],
     sales[type == 1],
     xlab='pre-sales',
     ylab='sales',
     pch=1,
     col='blue',
     xlim=c(min(pre.sales),max(pre.sales)),
     ylim=c(min(sales),max(sales)))
points(pre.sales[type == 2], sales[type == 2], pch=2, col='red')
points(pre.sales[type == 3], sales[type == 3], pch=15, col='green')

abline(intercept.1,slope,col='blue')
abline(intercept.2,slope,col='red')
abline(intercept.3,slope,col='green')
```