---
title: 'Discrete Multivariate Analysis - 579 - HW #5'
author: "Alex Towell (atowell@siue.edu)"
header-includes:
 - \usepackage{amsmath}
 - \usepackage{mathtools}
 - \usepackage{amsthm}
 - \usepackage{hhline}
output:
  pdf_document:
    df_print: kable
  html_document:
    df_print: paged
---

\newcommand{\var}{\operatorname{Var}}
\newcommand{\expect}{\operatorname{E}}
\newcommand{\corr}{\operatorname{Corr}}
\newcommand{\cov}{\operatorname{Cov}}
\newcommand{\ssr}{\operatorname{SSR}}
\newcommand{\se}{\operatorname{SE}}
\newcommand{\mat}[1]{\mathbf{#1}}
\newcommand{\RR}{\operatorname{RR}}
\newcommand{\eval}[2]{\left. #1 \right\vert_{#2}}

# Problem 1

Consider data from a prospective study on the relationship between daily
aspirin use and the onset of heart disease.

          Disease     No Disease     Total
 -------  -------     ----------     -----
 Placebo       28     656            684
 Aspirin       18     658            676

## Part (a)
\fbox{Provide an equation for $\sigma^2\!\left(\log \hat{\RR}\right)$.}

\begin{align*}
  \sigma^2\!\left(\log \hat{\RR}\right)
    &= \frac{1}{n_{1 1}} - \frac{1}{n_{1+}} + \frac{1}{n_{2 1}} - \frac{1}{n_{2+}}\\
    &= \frac{1}{28} - \frac{1}{684} + \frac{1}{18} - \frac{1}{658}\\
    &\approx 0.883\,.
\end{align*}


## Part (b)
\fbox{Provide an equation for $\sigma\!\left(\log \hat{\RR}\right)$.}
$$
   \sigma\!\left(\log \hat{\RR}\right) = \sqrt{\sigma^2(\log \hat{\RR})} \approx 0.297\,.
$$

## Part (c)
\fbox{Compute a $95\%$ confidence interval for $\log \RR$.}

The relative risk $\RR$ is defined as
$$
  \RR \coloneqq \frac{\pi_1}{\pi_2}
$$
The MLE of $\log \RR$ is given by
$$
  \hat{\RR} = \frac{\hat\pi_1}{\hat\pi_2} = \frac{n_{1 1}}{n_{1+}} \frac{n_{2+}}{n_{2 1}} = \frac{28}{676} \times \frac{684}{18} \approx 1.574\,.
$$
By the invariance property of the MLE, if $\hat\theta$ is an MLE of $\theta$
then $\operatorname{g}(\theta)$ is an MLE of $\operatorname{g}(\hat\theta)$.
Thus,
$$
  \log \hat{\RR} = \log \hat{\RR} \approx 0.454\,.
$$
A confidence interval for $\log \RR$ is thus
$$
  \operatorname{CI}(\log\RR) = \log \hat{\RR} \pm z_{\alpha/2} \sigma(\log \hat{\RR})\,.
$$
Letting $\alpha = 0.05$ and substituting in the values for $\log \hat{\RR}$ and $\sigma(\log \hat{\RR})$,
we get the result
$$
  \operatorname{CI}(\log\hat{\RR}) = 0.454 \pm  1.96 \times 0.297= [-0.128,1.036]\,.
$$

## Part (d)
\fbox{Compute a $95\%$ confidence interval for $\RR$.}

We take the inverse of the logarithm and obtain the result
$$
  \operatorname{CI}(\RR) = [e^{-0.128},e^{1.036}] = [0.880,2.818]\,.
$$


# Problem 2
A diagnostic test for Covid antibodies is being studied. For a sample of $n = 122$
specimens with antibodies known to be present, the test returned $y = 103$
positive results. For a sample of $n = 401$ specimens absent any antibodies,
the test returned $y = 399$ negative results.

## Function used to compute interval estimates for this problem.
```{r}
# this program performs Bayesian inference for a binomial probability
#   - y is the number of successes
#   - n is the number of trials
#   - alpha is the sum of the tail probabilities
#
# we also compute a classical interval using normality assumptions of the mle.
binomial_bayesian <- function(y,n,alpha=.05) {
  # compute the maximum likelihood estimator
  mle = y/n
  mle_var = mle*(1-mle)/n
  mle_sd = sqrt(mle_var)
  
  t0 = qt(1-alpha/2, n-1)
  L = mle - mle_sd*t0
  U = mle + mle_sd*t0
  
  # define the beta distribution parameters
  a=y+1
  b=n-y+1
  
  # computing a confidence interval by taking the upper and lower percentiles of
  # the beta distribution computing the median to represent the center of the
  # distribution
  lower = qbeta(alpha/2,a,b)
  median = qbeta(.5,a,b)
  upper = qbeta(1-alpha/2,a,b)
  cat("bayesian estimate: ", c(lower,median,upper),"\n")
  cat("mle confidence interval estimate: ", c(L,mle,U),"\n")
  
  # create a grid of p for plotting. you may change to a smaller range than from
  # 0 to 1 to see the distribution better
  p = seq(from=max(0,lower*.965),to=min(1,1.035*upper),length.out=100)
  # we are computing the beta density at each point in the grid
  posterior = dbeta(p,a,b)
  
  ps = rnorm(n=1000000,mean=mle,sd=mle_sd)
  
  # plotting the beta density, higher values of the curve represent stronger data
  # evidence
  plot(p,posterior,type = "l",col="blue")
  
  lines(density(ps),col="orange")
  
  points(median,0,col="blue",pch="*")
  points(c(lower,upper),c(0,0),,col="blue",pch="|")
  
  # let's plot the mle and confidence interval using t-score
  points(mle,0,col="orange",pch="*")
  points(c(L,U),c(0,0),,col="orange",pch="|")
}
```

## Part (a)
\fbox{Compute a Bayesian/likelihood interval estimate for the sensitivity,
$\delta = \Pr(+|A)$.}

Given that antibodies are present, the probability that a positive test result
is observed is given by the sensitivity $\delta$.

In classical statistics, while the probability $\delta$ is not known, the MLE
estimator is $\hat\delta = \frac{y}{n}$ which is asymptotically normal under
regularity conditions.

We compare it with the Bayesian approach, which does not assume normality.
The following R function call computes the Bayesian interval estimate for a
binomial probability. For comparison, we also compute the confidence interval
for the MLE using classical techniques that rely upon the normality assumption
of the MLE.

```{r}
binomial_bayesian(y=103,n=122,alpha=.05)
```

The maximum likelihood estimator $\hat\delta$ is $0.844$ with an $\alpha=0.05$
confidence interval $[0.779,0.909]$.

The point estimator of $\delta$ using the Bayesian approach, the $50\%$-percentile,
is $0.841$ with an interval estimate $[0.769,0.898]$.

The maximum likelihood estimator and its corresponding density plot and
confidence interval is plotted in \emph{orange} and the Bayesian estimator is
plotted in \emph{blue}.

## Part (b)
\fbox{Compute a Bayesian/likelihood interval estimate for the sensitivity,
$\gamma = \Pr(+|A)$.}

We do the same thing as in part (A).

```{r}
binomial_bayesian(y=399,n=401,alpha=.05)
```

The maximum likelihood estimator $\hat\delta$ is $0.844$ with an $\alpha=0.05$
confidence interval $[0.982,1.002]$

The point estimator of $\delta$ using the Bayesian approach, the $50\%$-percentile,
is $0.993$ with a $\alpha=0.05$ interval estimate $[0.982,0.998]$.

The maximum likelihood estimator and its corresponding density plot and
confidence interval is plotted in \emph{orange} and the Bayesian estimator is
plotted \emph{blue}.

# Problem 3
\fbox{
\begin{minipage}{.8\textwidth}
The clinical trial for the Moderna vaccine resulted in $n_I = 196$ cases of
symptomatic infection, with $y_C = 185$ cases from the control group, and
$y_V = 11$ cases from the vaccine group.
Compute a Bayesian/likelihood interval estimate for the vaccine efficacy,
$\operatorname{VE}$.
\end{minipage}}

The interval estimate is computed by the following R code.

```{r}
# this program performs a Bayesian inference for vaccine efficacy
# enter the number of infections in the vaccine group
# and the number of infections in the control group
v = 11
c = 185

# define the beta distribution parameters
a = v+1
b = c+1

# simulate a very large number of draws from the posterior distribution on p=P(V|I)
# for each simulated p, compute the value for vaccine efficacy VE
p = rbeta(1000000,a,b)
efficacy = 1-p/(1-p)

# create a plot of the posterior distribution on VE
hist(efficacy,xlab="Efficacy",probability = TRUE)
points(density(efficacy),type = 'l')

pt_est = quantile(efficacy,c(.5))
points(pt_est,0,col="blue",pch=8)

# compute percentiles for a 95% interval estimate
interval_est = quantile(efficacy,c(.025,.975))
points(interval_est,c(0,0),col="red",pch="|")

print(pt_est)
print(interval_est)
```

We see that the density plot is not quite normal (e.g., non-symmetric).
On the plot, we plotted the $50\%$ in blue and the $95\%$ interval estimate
in red.