---
title: 'STAT 581 - HW #2'
author: "Alex Towell (atowell@siue.edu)"
output:
  #md_document:
  #  variant: markdown_github
  pdf_document:
    df_print: kable
    latex_engine: xelatex
    #keep_tex: true
    highlight: tango
    #highlight: breezedark
    #highlight: pygments
    #highlight: zenburn
    #highlight: haddock
    #highlight: kate
    #highlight: espresso
fontsize: 11pt
geometry: margin=.5in
documentclass: article
#documentclass: paper
#documentclass: standalone
#classoption: twocolumn
header-includes:
 - \usepackage{amsmath}
 - \usepackage{mathtools}
 - \usepackage{amsthm}
 # - \usepackage{amsbsy}
 - \usepackage{bm}
 - \usepackage{xcolor}
 - \usepackage{fbox}
 - \usepackage{amsmath}
 - \usepackage{hw4}
---

```{r echo=F}
#knitr::opts_chunk$set(tidy.opts = list(width.cutoff = 80), tidy = TRUE)
knitr::opts_chunk$set(tidy.opts = list(width.cutoff = 80))
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

A product developer is investigating the tensile strength of a new synthetic
fiber that will be used to make cloth for men's shirts.
Strength may be affected by the percentage of cotton used in the blend of
material for the fiber.
A completely randomized experiment with five levels of cotton content is
performed.

# Part 1
\fbox{State the statistical hypothesis of interest.}

The hypothesis of interest is whether the percentage of cotton at five
different levels effects the tensile strength of a new synthetic fiber.

We may formulate this as a hypothesis test of the form
\begin{align*}
  H_0 &: \mu_1 = \cdots = \mu_5\\
  H_A &: \text{$\mu_i \neq \mu_j$ for at least one pair $(i,j)$, $i\neq j$,}
\end{align*}
where $\mu_k$ is the expected tensile strength at the $k$-th level of cotton.

If $H_0$ is true, the percentage of cotton has no effect on tensile strength.

# Part 2
\fbox{Briefly explain how the form of the alternative hypothesis requires a need
for further investigation.}

If there are differences in the cotton level means, further investigation is
required to determine where the differences occur.

# Part 3
\fbox{Create a Boxplot as a graphical display of the data.}

```{r}
library(printr)
library("readxl")

h2.data = read_excel("./handout2data.xlsx")
strength = na.omit(h2.data$strength)
percent = na.omit(as.factor(h2.data$percent))

boxplot(strength~percent)
```

# Part 4
\fbox{Compute the sample mean and sample variance of tensile strength for each level of cotton percentage.}

```{r}
means = by(strength,percent,mean)
variances = by(strength,percent,var)
cbind(means,variances)
```

# Part 5
\fbox{State the ANOVA model using treatment level effects. Compute estimates of the model parameters.}

In this CRD experiment, we observe $n=5$ responses at each of $a=5$ levels of
cotton (we treat the cotton percentage level as categorical, even though if we
show that the cotton level has a practical effect on tensile strength, we may
treat it as a quantitative input in, say, a regression model).

The data is given by
\begin{align*}
  Y_{1 1},\ldots,Y_{1 5} &\sim \mathcal{N}(\mu+\tau_1,\sigma^2)\\
  &\vdots\\
  Y_{5 1},\ldots,Y_{5 5} &\sim \mathcal{N}(\mu+\tau_5,\sigma^2).
\end{align*}

The model is given by
$$
  Y_{i j} = \mu + \tau_i + \epsilon_{i j}
  \begin{cases}
    i = 1,\ldots,5\\
    j = 1,\ldots,5,
  \end{cases}
$$
where
\begin{align*}
  Y_{i j}               &\qquad\text{is the $j$-th response for tensile strength for the $i$-th cotton level},\\
  \mu                   &\qquad\text{is the overall mean of the tensile strength},\\
  \tau_i                &\qquad\text{is the effect that the $i$-th cotton level has on the tensile strength},\\
  \epsilon_{i j}        &\qquad\overset{\rm{iid}}{\sim} \mathcal{N}(0,\sigma^2),\\
  \sum_{i=1}^{5} \tau_i &\qquad= 0.
\end{align*}

The estimates of the model parameters are given by
\begin{align*}
  \hat{\mu}       &= \bar{y}_{\cdot \cdot},\\
  \hat{\tau}_i    &= \bar{y}_{i \cdot} - \bar{y}_{\cdot \cdot}.
\end{align*}

We compute these estimates with the following R code:
```{r}
mu.hat = mean(strength)
tau.hat = means - mu.hat
```

We see that $\hat{\mu} = `r mu.hat`$ and
```{r test,echo=F}
df = data.frame(c(15,20,25,30,35),
                c(tau.hat[[1]],tau.hat[[2]],tau.hat[[3]],tau.hat[[4]],tau.hat[[5]]))
names(df) = c("cotton level", "$\\hat{\\tau}$")
knitr::kable(df)
```

# Part 6

## Part (a)
\fbox{\begin{minipage}{\textwidth}
Compute the $F_0$ statistic and the $p$-value.
Perform the statistical test at level $\alpha = .05$.
Provide an interpretation, stated in the context of the problem.
\end{minipage}}

```{r}
model.2 = aov(strength~percent)
summary(model.2)
```

We see that $F_0 = 14.76$ and the $p$-value $= .000$.
Thus, since $p = .000 < \alpha = .05$, the null model where cotton level
has no effect on tensile strength is not compatible with the data.

## Part (b)
\fbox{\begin{minipage}{\textwidth}
Compute the $t_0$ statistic and the $p$-value for testing the $30\%$ group
versus the $25\%$ group.
Provide an interpretation, stated in the context of the problem.
\end{minipage}}

All pair-wise hypothesis tests
\begin{align*}
  H_0^{(i,j)} &: \tau_i = \tau_j\\
  H_A^{(i,j)} &: \tau_i \neq \tau_j.
\end{align*}
may be computed with the following R code:
```{r}
# to perform multiple comparisons between groups,
# we will use the multcomp package
library("multcomp")
model.2.lsd = glht(model.2, linfct=mcp(percent = "Tukey"))
summary(model.2.lsd, test=univariate())
```

We are interested in the hypothesis test
\begin{align*}
  H_0^{(4,3)} &: \tau_4 = \tau_3\\
  H_A^{(4,3)} &: \tau_4 \neq \tau_3,
\end{align*}
which computes the results
$$
  t_0^{(4,3)} = 2.228
$$
and
$$
  p^{(4,3)} = .038.
$$

At $\alpha = .05$, this result is statistically significant.
Since $t_0^{(4,3)} < 0$, we decide $\tau_4 > \tau_3$.
The experiment finds that cotton level $4$ ($30\%$) has a larger effect on
tensile strength than cotton level $3$ ($25\%$).

## Part (c)
\fbox{\begin{minipage}{\textwidth}
Compute the $t_0$ statistic and the $p$-value for testing the $25\%$ group
versus the $20\%$ group.
Provide an interpretation, stated in the context of the problem.
\end{minipage}}

We are interested in the hypothesis test
\begin{align*}
  H_0^{(3,2)} &: \tau_3 = \tau_2\\
  H_A^{(3,2)} &: \tau_3 \neq \tau_2,
\end{align*}
which computes the results
$$
  t_0^{(3,2)} = 1.225
$$
and
$$
  p^{(3,2)} = .235.
$$

At $\alpha = .05$, this result is not statistically significant.
The experiment finds that cotton level $3$ ($25\%$) and cotton level $2$ ($25\%$)
have the same effect on tensile strength.

## Part (d)
\fbox{\begin{minipage}{\textwidth}
Use the grouping information for Fisher pairwise comparisons to further
investigate the nature of the treatment effect. Provide an interpretation,
stated in the context of the problem.
\end{minipage}}

```{r}
cld(summary(model.2.lsd, test=univariate()))
```

The experiment finds that $30\%$ cotton has the greatest effect on tensile
strength, $25\%$ and $20\%$ have the second greatest effect (but no difference
between them), and $15\%$ and $35\%$ have the weakest effect (but no difference
between them).

# Part 7
\fbox{\begin{minipage}{\textwidth}
Briefly discuss some other issues that may provide additional insight to an experimental result beyond
a finding of statistical significance.
\end{minipage}}

Statistical significance is a statement of evidence, not belief.
Other issues to consider:

1. The scientific context of the effect being tested.
   Is there an engineering explanation that may account for a relation between
   tensile strength and cotton level?

   Are there favorable past experimental effects? For instance, in the Bayesian
   perspective, we may encode results from prior experiments as a prior.

2. The complexity of the effect being tested, i.e., is multiple testing required
   to understand the effect?

3. The size of the effect, i.e., is the effect of practical importance? For
   instance, the data may not be compatible with the no effects model, but if
   the strength is only marginally effected by the cotton level, it may not be
   worth acting on this finding of statistical signifance.

4. The quality of the experimental design, i.e., is the type II error
   considered? Do we adjust for differences in experimental units or other
   factor effects?
