---
title: 'Time Series Analysis - STAT 478 - Project'
author: "Alex Towell (atowell@siue.edu)"
output:
  pdf_document:
    df_print: kable
    toc: true
    toc_depth: 2
    #latex_engine: pdflatex
    latex_engine: xelatex
header-includes:
 - \usepackage{amsmath}
 - \usepackage{mathtools}
 - \usepackage{amsthm}
 # - \usepackage{amsbsy}
 - \usepackage{bm}
 - \usepackage{xcolor}
---
\newcommand{\var}{\operatorname{Var}}
\newcommand{\expect}{\operatorname{E}}
\newcommand{\corr}{\operatorname{Corr}}
\newcommand{\cov}{\operatorname{Cov}}
\newcommand{\ssr}{\operatorname{SSR}}
\newcommand{\se}{\operatorname{SE}}
\newcommand{\mat}[1]{\bm{#1}}
\newcommand{\eval}[2]{\left. #1 \right\vert_{#2}}
\newcommand{\argmin}{\operatorname{arg\,min}}

# Confidentiality
We load the confidentiality measure as a time series $\{C_t\}$.
```{r}
library(dplyr)
data <- read.table(file="./acc1")
```

Let's look at a bit of the data:
```{r}
head(data)
tail(data)
```



```{r}
Ct <- c(1,2,3,4,5.1)
# Ct <- Ct / max(Ct)
```

Let's plot the time series:
```{r}
plot(Ct)
acf(Ct)
pacf(Ct)
```

This is non-stationary.
Variance is increasing as confidentiality increases, and the mean is non-constant.
In particular, there are a number of level shifts.
We'll try the log-transformation $\{ \log C_t \}$ to mitigate the non-constant
variance and then a first-order difference $\{\nabla \log C_t\}$:

```{r}
Ct.log <- log(Ct)
```

Let's plot it:
```{r}
plot(Ct.log)
```

This looks better, but of course the level shifts remain.

Let's take the difference to remove the level shifts, except for outliers
representing the level jumps.
Note that we must not lose the information on the level jumps.
They may seem like outliers in the difference plot, but they are one of the
most critical features.

```{r}
Ct.log.diff <- diff(Ct.log,1)
plot(Ct.log.diff)
```

The variance here seems reasonably constant.



```{r}
par(mfrow=c(1,2))
acf(Ct.log.diff)
pacf(Ct.log.diff)
```

Naturally, the level shifts are causing the spikes in the difference process.

The confidence interval is too small to see in the ACF and PACF.
Let's try to get a better look.

```{r}
acf(Ct.log.diff,ylim=c(-0.025, 0.025))
```

It seems quite a few values are outside of the confidence interval.

Let's do some residual plots to check for normality.
```{r}
library(forecast)
model <- arima(Ct.log.diff, c(0,0,1))
hist(model$residual,xlab="Residual",main="Histogram")
qqnorm(model$residual,main="QQ-plot")
acf(model$residual,main="Residuals")
pacf(model$residual,main="Residuals")
```

These aren't normally distributed residuals.

I think this calls for another modeling approach.
I saw that there are some R libraries that deal with level shifts.
Let's try \emph{changepoint}:

```{r include=FALSE}
library(changepoint)
#Ct <- Ct[5000:length(Ct)]
cp <- changepoint::cpt.meanvar(Ct,method="BinSeg", Q=26, minseglen=50)

plot(cp)

pts <- cpts(cp)
l <- 1
r <- -1

par(mfrow=c(1,2))
for (i in 2:length(pts))
{
  r <- pts[i]
  ys_seg <- Ct[l:r]
  ts_seg <- l:r
  ps_seg <- cbind(ts_seg,ys_seg)
  plot(ps_seg)
  model <- auto.arima(ys_seg)
  
  plot(model$residuals)

  l <- r+1
}
```


An interesting use-case for analying this time series data is not to forecast
from a particular data set, but to take the information in one or more time
series, deriving the distribution -- say, the distribution of changepoints and
their respective changes, and then generating plausible new time series
for predicting how potentially compromised one's confidentiality is.

In what follows, we consider the case of taking the change point data:
```{r include=FALSE}
ts <- append(pts,pts[length(pts)]+1000)
ys <- Ct[ts]
ps <- cbind(ts,ys)
plot(ps)
```

The change points actually seem pretty consistent.
Maybe it is uniformly distributed.
The question might then be, one is the expected jump with respect to the
change point.
Seems to follow a somewhat exponential curve, or at least some quadratic.
However, I only have the data from a single simulation. I need to fix up
my code and then run some more simulations if I want to pursue this angle
more, because certainly it may be the case that the curve changes significantly
from one run to the next.