---
title: 'Computational Statistics - STAT 575 - HW #2 - Covariance Estimation'
author: "Alex Towell (atowell@siue.edu)"
output:
  pdf_document:
    df_print: kable
    latex_engine: xelatex
header-includes:
 - \usepackage{amsmath}
 - \usepackage{mathtools}
 - \usepackage{amsthm}
 # - \usepackage{amsbsy}
 - \usepackage{booktabs}
 - \usepackage{bm}
 - \usepackage{xcolor}
 - \usepackage{fbox}
 - \usepackage{amsmath}
 - \usepackage{minted}
---

\newcommand{\var}{\operatorname{Var}}
\newcommand{\expect}{\operatorname{E}}
\newcommand{\cov}{\operatorname{Cov}}
\newcommand{\se}{\operatorname{SE}}
\newcommand{\mat}[1]{\mathbf{#1}} 
\newcommand{\eval}[2]{\left. #1 \right\vert_{#2}}
\newcommand{\poi}{\operatorname{POI}}
\newcommand{\argmax}{\operatorname{arg\,max}}

# Covariance estimation: bootstrap vs observed fisher information

We use the functions defined by the package \emph{prob.4.2.comp.stats}.

```{r message=F}
if (!require(devtools)) { install.packages('devtools') }
install_github("queelius/prob.4.2.comp.stats")
library("prob.4.2.comp.stats")
options(tinytex.engine_args = '-shell-escape')
```

We invoke the EM algorithm on the observed count data.
```{r}
counts <- c(379,299,222,145,109,95,73,59,45,30,24,12,4,2,0,1,1)

# theta is initial guess for parameter vector,
#   theta = (alpha, beta, mu, lambda)' = (1/3,1/3,2,3).
# the initial value assumes each category is equally probable and mu^(0) and
# lambda^(0) are just arbitrarily chosen to be 2 and 3, with the insight that
# group 3 is more risky than group 2.
theta <- c(1/3,1/3,2,3)
sol <- em.estimator(theta,counts)
theta.em <- sol$estimate
```

We see that $\hat{\theta}_{\rm{EM}} = (`r theta.em`)'$.
We use the Bootstrap method to estimate the covariance matrix of
$\hat{\theta}_{\rm{EM}}$ in the following:
```{r}
cov.bs <- em.cov.bs(theta.em,counts,m=10000)
knitr::kable(cov.bs)
```

For comparison, we use the observed information matrix, the Hessian of the
observed log-likelihood evaluated at $\hat{\theta}_{\rm{EM}}$,
$$
  \hat{\Sigma} = \eval{\operatorname{H}(-\ell(\theta))}{\hat{\theta}_{\rm{EM}}}.
$$

```{r}
cov.ob <- em.cov.info(theta.em,counts)
knitr::kable(cov.ob)
```

We see that these covariance matrices appear to be different.
It is not clear why.
We have separated out the relevant code in the package and put it into the
following code block.
Dr. Q, do you see anything wrong with this?

```{r}
em.cov.info(theta.em,counts)
```