---
title: "Regression Analysis - STAT 482 - Exam 2: Due Dec 14, 2021"
author: "Alex Towell (atowell@siue.edu)"
geometry: margin=1cm
header-includes:
- \usepackage{amsmath}
- \usepackage{mathtools}
- \usepackage{amsthm}
- \usepackage{color}
- \usepackage{xcolor}
- \usepackage{geometry}
- \usepackage{enumerate}
output:
  pdf_document:
    toc: no
    latex_engine: xelatex
    #df_print: kable
editor_options:
  markdown:
    wrap: 80
---

```{=tex}
\newcommand{\sos}[1]{\mathrm{SS_{#1}}}
\newcommand{\ms}[1]{\mathrm{MS_{#1}}}
\newcommand{\sd}{\operatorname{sd}}
\newcommand{\var}{\operatorname{var}}
\newcommand{\expect}{\operatorname{E}}
\newcommand{\corr}{\operatorname{cor}}
\newcommand{\cov}{\operatorname{cov}}
\newcommand{\se}{\operatorname{se}}
\newcommand{\eval}[2]{\left. #1 \right\vert_{#2}}
\newcommand{\degf}[1]{\mathrm{df_{#1}}}
\newcommand{\entropy}{\operatorname{H}}
\newcommand{\param}[1]{\textcolor{blue}{\texttt{#1}}}
\newcommand{\ci}{\operatorname{CI}}
```

```{r, include=FALSE}
options(tinytex.engine_args = '-shell-escape')
```

# Problem 1
> A beer distributor is interested in the amount of time to service its retail
> outlets. Two factors are thought to influence the delivery time ($y$) in
> minutes: the number of cases delivered ($x_1$) and the distance traveled
> ($x_2$) in miles. A random sample of delivery time data has been collected.
> The data is available on Blackboard as a csv file.

## P1: (a)
> Provide an interpretation of a regression coefficient in a multiple regression
> model.

## P1: (b)
> Compute $b_1,b_2$, the estimated the regression coefficients for the delivery
> time data.

```{r}
data1 = read.csv('exam2-1.csv')
head(data1)
data1.m12 = lm(time ~ cases + distance, data=data1)
data1.m12
```

## P1: (c)
> Compute $t$ statistics for testing the effect of each input variable. Explain
> what type of effect is being tested here.

```{r}
coef(summary(data1.m12))
```

## P1: (d)
> Compute $\sos{R}(X_1)$ and $\sos{R}(X_2|X_1)$.
> Explain what each sum of squares represents.

```{r}
anova(data1.m12)
```

## P1: (e)
> Test for a marginal effect of $x_2$ against a model which includes no other
> input variables. (Compute the test statistic and $p$-value.) Provide an
> interpretation of the result, stated in the context of the problem.

```{r}
data1.m0 = lm(time ~ 1, data=data1)
data1.m2 = lm(time ~ distance, data=data1)
anova(data1.m2)
anova(data1.m0,data1.m2)  # equivalent to anova(data1.m2)
```

## P1: (f)
> Test for a partial effect of $x_2$ against a model which includes $x_1$.
> (Compute the test statistic and $p$-value.)
> Provide an interpretation of the result, stated in the context of the problem.

```{r}
data1.m1 = lm(time ~ cases, data=data1)
anova(data1.m1,data1.m12)
```

## P1: (g)
> Compute the correlation matrix. What feature of multidimensional modeling is
> illustrated in this problem?

```{r}
cor(data1)
```

# Problem 2
> A bakery is interested in the best formulation for a new product. A
> small-scale experiment is conducted to investigate the relationship between
> the product satisfaction ($y$), and the moisture content (input 1) and
> sweetness (input 2) of the product. The input variables have been coded
> $(x_1,x_2)$ for ease of calculation. The data is available on Blackboard as a
> csv file.

## P2: (c)
> Fit an interaction model using the coded variables. Compute the regression
> coefficient estimates and their standard errors.

```{r}
data2 = read.csv('exam2-2.csv')
data2
data2.xmod = lm(satisfaction ~ x1*x2, data=data2)
summary(data2.xmod)
```

## P2: (d)
> Write the estimated regression as a function of $x_1$ for $x_2 = 1,0,-1$.

```{r}
b0 = coef(data2.xmod)[1]
b1 = coef(data2.xmod)[2]
b2 = coef(data2.xmod)[3]
b12 = coef(data2.xmod)[4]
reg.est.x1 = matrix(c(b0+b2,b1+b12,  # x2=+1
                      b0,b1,         # x2=0
                      b0-b2,b1-b12), # x2=-1
                    nrow=3,byrow=T)

dimnames(reg.est.x1) = list(c("x2=+1","x2=0","x2=-1"),c("intercept","slope"))
reg.est.x1
```

## P2: (e)
> Create interaction plots for both the interaction model and the additive
> effects model.

```{r}
# interaction model:
data2.xmod.fits = predict(data2.xmod)
interaction.plot(data2$x1,data2$x2,data2.xmod.fits,
            col=c("red","blue"),
            trace.label="x2",
            xlab="x1",
            ylab="y")

# additive model
data2.amod = lm(satisfaction ~ x1+x2,data=data2)
data2.amod.fits = predict(data2.amod)
interaction.plot(data2$x1,data2$x2,data2.amod.fits,
            col=c("red","blue"),
            trace.label="x2",
            xlab="x1",
            ylab="y")
```

## P2: (f)
> Test for an interaction effect. (Compute the test statistic and $p$-value.)

```{r}
anova(data2.amod, data2.xmod)
```

# Problem 3
> An engineer is interested in comparing three chemical processes (categorical
> input with groups $A$,$B$,$C$) for manufacturing a compound. It is suspected
> that the impurity (continuous input $x$) of the raw material will affect the
> yield (response variable $y$) of the product. The data is available on
> Blackboard as a csv file.

## P3: (e)
> Compute interval estimates for each of the effect parameters.

```{r}
library(matlib)
data3 = read.csv('exam2-3.csv')
data3$process = as.factor(data3$process)

contrasts(data3$process) = contr.treatment(3,base=3)
data3.amod = lm(yield ~ impurity+process,data=data3)

b.hat = coef(data3.amod)
dfe = nrow(model.matrix(data3.amod)) - ncol(model.matrix(data3.amod))
V = vcov(data3.amod)
a = c(0,0,1,-1)

b.hat.12 = a %*% b.hat
se.12 = sqrt(a %*% V %*% a)

b.hat.12.lower = b.hat.12 - qt(.975,dfe) * se.12
b.hat.12.upper = b.hat.12 + qt(.975,dfe) * se.12

confint(data3.amod)

cat("beta2-beta1",c(b.hat.12.lower,b.hat.12.upper))
```

## P3: (f)
> Create a scatterplot of the data with the estimated regression lines.

```{r}
intercept.3 = b.hat[1]
intercept.1 = b.hat[1]+b.hat[3]
intercept.2 = b.hat[1]+b.hat[4]
slope = b.hat[2]

cat(slope, intercept.1, intercept.2, intercept.3)

attach(data3)

plot(impurity[process == "A"],yield[process == "A"],
     xlab='impurity',ylab='yield',pch=1,
     xlim=c(min(impurity)-1,max(impurity)+1),
     ylim=c(min(yield)-1,max(yield)+1))

points(impurity[process=="B"], yield[process=="B"], pch=2)
points(impurity[process=="C"], yield[process=="C"], pch=15)

abline(intercept.1,slope,lty=1)
abline(intercept.2,slope,lty=2)
abline(intercept.3,slope,lty=3)
```