---
title: 'Discrete Multivariate Analysis - 579 - HW #8'
author: "Alex Towell (atowell@siue.edu)"
header-includes:
 - \usepackage{amsmath}
 - \usepackage{mathtools}
 - \usepackage{amsthm}
 - \usepackage{multirow}
 - \usepackage{booktabs}
output:
  pdf_document:
    df_print: kable
  html_document:
    df_print: paged
---

\newcommand{\var}{\operatorname{Var}}
\newcommand{\expect}{\operatorname{E}}
\newcommand{\corr}{\operatorname{Corr}}
\newcommand{\cov}{\operatorname{Cov}}
\newcommand{\ssr}{\operatorname{SSR}}
\newcommand{\se}{\operatorname{SE}}
\newcommand{\mat}[1]{\mathbf{#1}}
\newcommand{\RR}{\operatorname{RR}}
\newcommand{\eval}[2]{\left. #1 \right\vert_{#2}}

# Problem 1
Consider an experiment on chlorophyll inheritance in maize.
A genetic theory predicts the ratio of green to yellow to be 3:1.
In a sample of $n = 1103$ seedlings, $n_1 = 854$ were green and $n_2 = 249$ were
yellow.

## Part (a)
\fbox{Compute the statistic $G^2$ for testing the proposed model.}

The likelihood ratio statistic is defined as
$$
  G^2 = - 2 \log \Lambda = 2 \sum_{j=1}^{c} \log\left(\frac{n_j}{n \pi_{j 0}}\right).
$$

We use the following R code to compute the observed statistic $G_0^2$.

```{r}
# below is code for testing a specified multinomial
# data is entered as a vector of counts
obs = c(854,249)
n = sum(obs)

# the null model is entered as a vector of hypothesized probabilities
# expected counts under the null model are computed as n*prob
pi.0 = c(.75,.25)
m = n*pi.0

# log-likelihood ratio test
G2 = 2*sum(obs*log(obs/m))
print(G2)
```
We see that the observed statistic is given by $G_0^2 = 3.54$.

## Part (b)
\fbox{Compute the $p$-value for the proposed model.}

The reference distribution is the chi-squared distribution with $c-1 = 1$ degrees
of freedom.

The $p$-value with $k$ degrees of freedom is defined as
$$
  \text{$p$-value} = \operatorname{P}(\chi^2(k) > G^2),
$$
or in this case
$$
  \text{$p$-value} = \operatorname{P}(\chi^2(1) > 3.54) = 0.06.
$$
The following R code computes the $p$-value.
```{r}
# the p-value, a calculation based on tail probabilities
c = length(obs)
pvalue.G2 = pchisq(G2,c-1,lower.tail=FALSE)
print(pvalue.G2)
```

## Part (c)
\fbox{Provide an interpretation of your result, stated in the context of the problem.}

According to the Fisher scale for interpreting $p$-values, $p$-values of $0.05$
and $0.1$ denote respectively \emph{moderate} and \emph{boderline} evidence
against the null model.

Since we obtained a $p$-value of $0.06$, the data provides \emph{moderate}
evidence against the genetic theory positing that the ratio of green to yellow
seedlings is 3:1.

## Part (d)
\fbox{What are the shortcomings of using a $p$-value as a measure of evidence?}

The $p$-value overstates evidence against the null model by comparing
the null model to the alternative model that is best supported by the data.

## Part (e)
\fbox{
\begin{minipage}{0.8\textwidth}
Compute the Bayes Factor $B_{0 1}$ for medium, wide, and ultrawide prior
distributions.
According to the Bayes Factors, which model is supported by the data?
\end{minipage}
}

Below is code for computing Bayes Factors in a test for a specified probability.
```{r message=FALSE}
library(BayesFactor)
# g denotes number of green seedlings
# y denotes number of yellow seedlings
# n = g+y denotes total number of seedlings
g = 854
y = 249
n = g+y
p0 = 3/4

bf.medium = 1/proportionBF(y=g,N=n,p=p0,rscale="medium")
bf.wide = 1/proportionBF(y=g,N=n,p=p0,rscale="wide")
bf.ultra = 1/proportionBF(y=g,N=n,p=p0,rscale="ultrawide")

bf.medium

bf.wide

bf.ultra
```

If $\rm{BF}_{0 1} > 1$, then the data supports the null model over the
alternative model.

According to the Bayes Factors, the results for \emph{medium}, \emph{wide} and
\emph{ultrawide} are given respectively by $1.9$, $2.7$, and $3.8$.
Thus, the null model is supported by the data in each case.