---
title: 'Regression Analysis - STAT 482 - HW #3'
author: "Alex Towell (atowell@siue.edu)"
header-includes:
 - \usepackage{amsmath}
 - \usepackage{mathtools}
 - \usepackage{amsthm}
 - \usepackage{multirow}
 - \usepackage{booktabs}
 - \usepackage{minted}
 - \usepackage{color}
 - \usepackage{xcolor}
 - \usepackage{tcolorbox} 
output:
  pdf_document:
    toc: false
    latex_engine: xelatex
    df_print: kable
  html_document:
    df_print: paged
---

\newsavebox{\selvestebox}
\newenvironment{colbox}[1]
  {\newcommand\colboxcolor{#1}%
   \begin{lrbox}{\selvestebox}%
   \begin{minipage}{\dimexpr\columnwidth-2\fboxsep\relax}}
  {\end{minipage}\end{lrbox}%
   \begin{center}
   \colorbox[HTML]{\colboxcolor}{\usebox{\selvestebox}}
   \end{center}}

\newcommand{\ssx}{\rm{SS}_x}
\newcommand{\var}{\operatorname{V}}
\newcommand{\sd}{\operatorname{sd}}
\newcommand{\expect}{\operatorname{E}}
\newcommand{\corr}{\operatorname{Corr}}
\newcommand{\cov}{\operatorname{Cov}}
\newcommand{\ssr}{\operatorname{ssr}}
\newcommand{\se}{\operatorname{se}}
\newcommand{\mat}[1]{\mathbf{#1}}
\newcommand{\eval}[2]{\left. #1 \right\vert_{#2}}

```{r, include=FALSE}
options(tinytex.engine_args = '-shell-escape')
```

Refer to the data from Exercise 1.20

Data has been collected on $45$ calls for routine maintenance.
The goal is to explore the relationship between the number of copiers serviced
$(x)$ and the time in minutes spent to complete the service $(y)$.
Let $x_h = 3$ copiers.

# Part (1)
\fbox{Compute a confidence interval for $\mu_h$.}

$\mu_h$ is the mean response $\expect(Y_h|x_h=3)$.
A confidence interval for $\mu_h$ is given by
$$
  \hat{Y}_h \pm t_{\alpha/2,n-2} \se(\hat{Y}_h).
$$

```{r}
alpha = .05
x.h = 3
data = read.table('CH01PR20.txt')
colnames(data)=c("time","copiers")
mod = lm(time~copiers, data=data)
predict(mod,level=1-alpha,data.frame(copiers=x.h),interval="confidence")
```

We see that a CI for $\mu_h$ is given by $[44.526,47.904]$.

# Part (2)
\fbox{Compute a prediction interval for $Y_{h(new)}$.}

We denote $Y|X=x_h$ by $Y_h$.
Assume
$$
  Y_h \sim \mathcal{N}(\beta_0 + \beta_1 x_h, \sigma^2).
$$
Then, if we wish to predict $Y_h$ with $(\beta_0,\beta_1,\sigma)$ known, the
interval
$$
  \beta_0 + \beta_1 x \pm z_{\alpha/2} \sigma.
$$
includes the observed value of $Y_h$ with probability $1-\alpha$.
For predicting a single outcome $Y_h$, the uncertainty of the prediction is
quantified by $\var(Y_h) = \sigma^2$.
For a given $\alpha$, the smaller the variance the smaller the interval.

However, if $(\beta_0,\beta_1,\sigma^2)'$ is unknown, we have the additional
source uncertainty due to a random sample $\{(Y_h,x_h)\}$ being used to
estimate $(\beta_0,\beta_1,\sigma^2)'$.

Let `pred` = $P_h = Y_{h(\rm{new})} - \hat{Y}_h$ denote the random \emph{prediction error}.
Then, $\expect(P_h) = 0$ and
\begin{align*}
  \var(P_h) &= \var(Y_{h(\rm{new})}) + \var(\hat{Y}_h)\\
            &= \sigma^2\left[1 + \frac{1}{n} + \frac{(x_h - \bar{x})^2}{\ssx}\right].
\end{align*}

Since we do not know $\sigma^2$, we estimate it with $\rm{MSE}$ and thus a
prediction interval for $Y_{h(\rm{new})}$ is given by
$$
  \hat{Y}_h \pm t_{\alpha/2,n-2} \hat{\sd}(P_h)
$$
where
$$
  \hat{\sd}(P_h) = \sqrt{\rm{MSE}\left(1 + \frac{1}{n} + \frac{(x_h-\bar{x})^2}{\ssx}\right)}.
$$

```{r}
predict(mod,data.frame(copiers=3),level=1-alpha,interval="predict")
```

We see that the PI for $Y_{h(\rm{new})}$ is given by $[26.235,62.816]$.

# Part (3)
\fbox{\begin{minipage}{\textwidth}
Explain the difference between a confidence interval and a prediction interval,
stated in the context of the problem.
\end{minipage}}

A confidence interval is for the mean service time for all service calls
with $3$ copiers to service.

A prediction interval is for the service time of a single service call
with $3$ copiers to service.

# Part (4)
\fbox{Let $m=10$. Compute a prediction interval for $\bar{Y}_{h(\rm{new})}$.}

Let `pred.mean` = $\bar{P}_h = \bar{Y}_{h(\rm{new})} - \hat{Y}_h$ denote the random \emph{prediction error}.
Then, $\expect(\bar{P}_h) = 0$ and
\begin{align*}
  \var(\bar{P}_h)
    &= \var(\bar{Y}_{h(\rm{new})}) + \var(\hat{Y}_h)\\
    &= \sigma^2 \left[\frac{1}{m} + \frac{1}{n} + \frac{(x_h - \bar{x})^2}{\ssx}\right].
\end{align*}

Since we do not know $\sigma^2$, we estimate it with $\rm{MSE}$ and thus a
prediction interval for $\bar{Y}_{h(\rm{new})}$ is given by
$$
  \hat{Y}_h \pm t_{\alpha/2,n-2} \hat{\sd}(\bar{P}_h)
$$
where
$$
  \hat{\sd}(\bar{P}_h) = \sqrt{\rm{MSE}\left(\frac{1}{m} + \frac{1}{n} + \frac{(x_h-\bar{x})^2}{\ssx}\right)}
$$


```{r}
b0 = mod$coefficients[1]; names(b0) = NULL
b1 = mod$coefficients[2]; names(b1) = NULL

residual = mod$residuals
n = length(residual)
sse = sum(residual^2)
mse = sse / (n-2)
x = data$copiers

x.bar = mean(x)
ssx = sum((x-x.bar)^2)

y.hat = b0 + b1*x.h 
m = 10
var.pred.mean = mse*(1/m + 1/n + (x.h-x.bar)^2/ssx)

lower.ynew = y.hat - qt(alpha/2,n-2,lower.tail=F)*sqrt(var.pred.mean)
upper.ynew = y.hat + qt(alpha/2,n-2,lower.tail=F)*sqrt(var.pred.mean)

c(lower.ynew,upper.ynew)
```

We see that the PI for $\bar{Y}_{h(\rm{new})}$ is given by
$$
  [`r round(lower.ynew,digits=3)`, `r round(upper.ynew,digits=3)`].
$$

# Part (5)
\fbox{Provide an interpretation of your result, stated in the context of the problem.}

We predict that the mean service time for $10$ service calls with $3$ copiers each
will be between $`r round(lower.ynew,digits=3)`$ and $`r round(upper.ynew,digits=3)`$.

# Part (6)
\fbox{\begin{minipage}{\textwidth}
Show that $\var(\rm{pred.mean})$ converges to $\var(\hat{Y}_h)$ as $m \to \infty$.
Use this limit result to interpret the confidence interval for $\mu_h$ as a prediction.
\end{minipage}}

The variance of `pred.mean` = $\bar{P}_h$ is given by
$$
  \var(\bar{P}_h) = \sigma^2 \left(\frac{1}{m} + \frac{1}{n} + \frac{(x_h-\bar{x})^2}{\ssx}\right)
$$
which we may rewrite as
\begin{align*}
  \var(\bar{P}_h)
    &= \frac{\sigma^2}{m} + \sigma^2\left(\frac{1}{n} + \frac{(x_h-\bar{x})^2}{\ssx}\right)\\
    &= \var(\bar{Y}_{h(\rm{new})}) + \var(\hat{Y}_h).
\end{align*}

As $m \to \infty$, $\var(\bar{Y}_{h(\rm{new})}) \to 0$, and thus
$$
  \lim_{m \to \infty} \var(\bar{P}_h) = \var(\hat{Y}_h).
$$

A CI estimate of the mean service time $\mu_h$ can be interpreted as a
prediction of the sample mean $\bar{Y}_{h(\rm{new})}$ for a large number of
service time responses.