---
title: 'STAT 581 - HW #3'
author: "Alex Towell (atowell@siue.edu)"
output:
  #md_document:
  #  variant: markdown_github
  pdf_document:
    df_print: kable
    latex_engine: xelatex
    #keep_tex: true
    highlight: tango
    #highlight: breezedark
    #highlight: pygments
    #highlight: zenburn
    #highlight: haddock
    #highlight: kate
    #highlight: espresso
fontsize: 11pt
geometry: margin=.5in
documentclass: article
#documentclass: paper
#documentclass: standalone
#classoption: twocolumn
header-includes:
 - \usepackage{amsmath}
 - \usepackage{mathtools}
 - \usepackage{amsthm}
 # - \usepackage{amsbsy}
 - \usepackage{bm}
 - \usepackage{xcolor}
 - \usepackage{fbox}
 - \usepackage{amsmath}
 - \usepackage{hw4}
---

```{r echo=F}
#knitr::opts_chunk$set(tidy.opts = list(width.cutoff = 80), tidy = TRUE)
knitr::opts_chunk$set(tidy.opts = list(width.cutoff = 80))
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

# Problem 1

An experiment is conducted to study the effect of drilling method on drilling
time.
Each method (dry drilling, wet drilling) is used on $n = 12$ rocks.
Drilling times are measured in $1/100$ minutes.

## Part (a)
\fbox{\begin{minipage}{\textwidth}
Compute a $95\%$ confidence interval for $\delta = \mu_1 - \mu_2$.
Provide an interpretation, stated in the context of the problem.
\end{minipage}}

A confidence interval for $\delta$ includes all parameter values compatible with
the observed data $\hat\delta$,
$$
  \delta \in \left[
    \hat\delta + t_{\alpha/2,2(n-1)} s_p \sqrt{2/n},
    \hat\delta + t_{1-\alpha/2,2(n-1)} s_p \sqrt{2/n}
  \right].
$$

We compute this CI with:
```{r}
library("readxl")
data = read_excel("./handout1data.xlsx")
data$method = as.factor(data$method)
dry = na.omit(data$time[data$method=='d'])
wet = na.omit(data$time[data$method!='d'])
alpha = .05

t.test(x=dry,
       y=wet,
       alternative=c("two.sided"),
       conf.level=1-alpha,
       var.equal=T)$conf.int[1:2]
```

We estimate that the difference in drilling methods,
$\delta = (\rm{dry} - \rm{wet})$, is between $[126.876,276.458]$.

## Part (b)
\fbox{\begin{minipage}{\textwidth}
Explain how a confidence interval provides a complementary result to a
hypothesis test.
\end{minipage}}

A hypothesis test looks to determine if an effect exists.
A CI looks to determine the size of the effect.

# Problem 2
A prodcut developer is investigating the tensile strength of a new synthetic
fiber.
A completely randomized design with five levels of cotton content is performed,
with $n=5$ speciments per level.

## Part (a)
\fbox{\begin{minipage}{\textwidth}
Compute and display $95\%$ confidence intervals for all pairwise comparisons.
\end{minipage}}

We show the confidence intervals with:
```{r}
library("multcomp")
data.2 = read_excel("./handout2data.xlsx")
strength = na.omit(data.2$strength)
percent = na.omit(as.factor(data.2$percent))

m = aov(strength~percent)
m.lsd = glht(m,linfct=mcp(percent="Tukey"))
ci.lsd = confint(m.lsd,calpha=univariate_calpha())
plot(ci.lsd)
ci.lsd
```

## Part (b)
\fbox{\begin{minipage}{\textwidth}
Explain how a confidence interval can be used in testing $H_0 : \mu_i = \mu_j$,
for each pair of factor levels.
\end{minipage}}

If the CI for $\mu_i - \mu_j$ contains $0$, the decision is
to decide $H_0^{(i,j)} : \mu_i - \mu_j$.
Otherwise, the decision is to decide $H_A^{(i,j)} : \mu_i \neq \mu_j$.

## Part (c)
\fbox{\begin{minipage}{\textwidth}
Explain how computing multiple intervals impacts the probability of committing an error.
\end{minipage}}

Suppose the hypothesis is given by
\begin{align*}
  H_0 &: \mu_1 = \cdots = \mu_a\\
  H_A &: \mu_i \neq \mu_j \, \text{for at least one pair $(i,j)$, $i\neq j$.}
\end{align*}

If we approximate this hypothesis test with $\binom{a}{2}$ pairwise tests of the
form
\begin{align*}
  H_0^{(i,j)} &: \mu_j = \mu_i\\
  H_0^{(i,j)} &: \mu_j \neq \mu_i
\end{align*}
for $i=1,\ldots,a-1$ and $j=i+1,\ldots,a$ where
$$
  \Pr\{\text{decide} \, H_A^{(i,j)} | H_0^{(i,j)} \, \text{true}\} = \alpha,
$$
then a type I error occurs if one or more of the tests is a false positive.

We denote this false positive rate by $\alpha_g$ which satisfies
\begin{align*}
  \alpha_g &= \Pr\{\text{decide} \, H_A | H_0 \, \text{true}\}\\
           &= \Pr\{\text{decide} \, H_A^{(i,j)} \, \text{for some $(i,j)$} | H_0 \, \text{true}\}\\
           &> \alpha.
\end{align*}

\begin{proof}
The equation for $\alpha_g$ may be rewritten as
$$
  \alpha_g = 1-\Pr\{\text{decide} \, H_0^{(i,j)} \, \text{for all} \, (i,j) | H_0 \, \text{true}\}.
$$

These pairwise tests are not necessarily independent, so assume $k$ independent
tests, $1 < k \leq \binom{a}{2}$, each with false positive rate $\alpha$.
The probability that a false positive does not occur on a pairwise test is $1-\alpha$,
so we rewrite the above as
$$
  \alpha_g = 1-(1-\alpha)^k,
$$
which satisfies $\alpha_g > \alpha$ over its support $(0,1)$.
\end{proof}

Observe that $\alpha_g \to 1$ as $k \to \infty$ and $\alpha_g = \alpha$ at $k=1$.

# Problem 3
An experiment to compare a new drug to a standard is in the planning stages.
The response variable of interest is the clotting time (in minutes) of blood
drawn from a subject.
The experimenters want to perform a two sample $t$ test at level $\alpha = .05$,
having power $\pi = .8$ at $\delta_A = 0.25$, for standard deviation $\sigma=0.7$.

## Part (a)
\fbox{\begin{minipage}{\textwidth}
Determine the sample size for each drug in order to achieve the stated test
specifications.
\end{minipage}}

We compute $n$ with:
```{r}
sd = .7
alpha = .05
h = .8  # power
v = .25 # alternative
power.t.test(n=NULL,delta=v,sd=sd,sig.level=alpha,power=h,type="two.sample")
```

We see that $n' = 124.0381$.
If the power $h$ is a lower-bound, then
let $n=\lceil n' \rceil = 125$.
If the power specification is not a lower-bound, but at approximate specification,
it may be appropriate to round to the nearest integer, $n=[ n' ] = 124$.

## Part (b)
\fbox{\begin{minipage}{\textwidth}
Graph the power curve for the chosen sample size.
Explain how the power curve displays the desired properties of the test.
\end{minipage}}

```{r}
# power.curve
#
# create the power curve for the chosen sample size.
#
# arguments:
#   n: sample size,
#   sd: standard deviation
#   alpha: significance level
#   h: power (shows as a horizontal line)
#   v: specific alternative (shows as a vertical line)
#
# output:
#   graph of power curve
power.curve = function(n, sd, alpha, h, v)
{
  df = 2*(n-1)
  delta = seq(from=0,to=5*sd/sqrt(n/2),length.out = 1000)
  power = 1 - pt(qt(1-alpha/2,df),df,ncp = sqrt(n/2)*(delta/sd))
  
  plot(delta,power,type = "l",lwd=2,col="blue")
  abline(h=h,col="red",lwd=2)  
  abline(v=v,col="green",lwd=2)
}

power.curve(n=124,sd=sd,alpha=alpha,h=h,v=v)
```

First, we see that the power at $\delta_A = `r v`$ obtains a power of
$\pi(\delta_A) = 0.8$, as required.

Second, and this is not explicitly shown on the graph,
whether $H_0 : \delta = 0$ is true or $H_A : \delta = \delta_A$ is true, there is a low probability of committing an error
since the $95\%$ confidence interval under the null model, approximately
$$
  \pm 1.96 \sigma / \sqrt{n} = [`r round(-1.96*sd/sqrt(124),digits=3)`,`r round(1.96*sd/sqrt(124),digits=3)`],
$$
does not intersect with the $95\%$ confidence interval under the alternative model,
$$
  \delta_A \pm 1.96 \sigma / \sqrt{n} = [`r round(v - 1.96*sd/sqrt(124),digits=3)`,`r round(v + 1.96*sd/sqrt(124),digits=3)`].
$$

The small region between these confidence intervals may be classified
as the "don't care" region.

## Part (c)
\fbox{\begin{minipage}{\textwidth}
Provide a general explanation of how $\delta_A$ can be determined.
\end{minipage}}

The specific alternative $\delta_A$ is chosen to represent an effect size
that is expected (e.g., from past experience or related data), important
(difference is non-negligible), and/or practical (cost considerations).

# Problem 4
Refer back to the tensile strength exampale of problem 2.
Use the data from this study to perform a power analysis for a main study.
The experimenters desire a level $\alpha = .05$ test with power $\pi = .8$.

## Part (a)
\fbox{\begin{minipage}{\textwidth}
Determine the sample size for each group based on specifying the maximum
difference in means.
\end{minipage}}

```{r}
pwr = .8
alpha = .05
a = 5
means = by(strength,percent,mean)
max.D = max(means) - min(means)
#summary(aov(strength~percent))
s2 = 8.06
power.anova.test(
  groups=a,
  between.var=max.D^2/2/(a-1),
  within.var=s2,
  power=pwr,
  sig.level=alpha,
  n=NULL)
```

We see that $n' = 2.534$.
Rounding to the nearest integer, we propose using a sample size of $n=3$ for each group.

## Part (b)
\fbox{\begin{minipage}{\textwidth}
Use a simulation to compute power at $n=3$ using the pilot study to specify
the model parameters.
\end{minipage}}

```{r}
sim.size = 10
decide.Ha = rep(NA,sim.size)
n = 3

for (k in 1:sim.size)
{
  y1 = rnorm(n,means[1],sqrt(s2))
  y2 = rnorm(n,means[2],sqrt(s2))
  y3 = rnorm(n,means[3],sqrt(s2))
  y4 = rnorm(n,means[4],sqrt(s2))
  y5 = rnorm(n,means[5],sqrt(s2))
  ybar1 = mean(y1)
  ybar2 = mean(y2)
  ybar3 = mean(y3)
  ybar4 = mean(y4)
  ybar5 = mean(y5)
  var1 = var(y1)
  var2 = var(y2)
  var3 = var(y3)
  var4 = var(y4)
  var5 = var(y5)
  F.stat = n*var(c(ybar1,ybar2,ybar3,ybar4,ybar5)) / mean(c(var1,var2,var3,var4,var5))
  decide.Ha[k] = (F.stat>qf(1-alpha,a-1,a*(n-1)))
}
power = mean(decide.Ha)
power
```


## Part (c)
\fbox{\begin{minipage}{\textwidth}
Comment on the use of pilot study data in a power analysis.
\end{minipage}}

Specifying parameter values for a power analysis based on estimates from a pilot
study, without accounting for estimation error, may lead to a hypothesis test
that does not have adequate power.