---
title: 'Regression Analysis - STAT 482 - HW #2'
author: "Alex Towell (atowell@siue.edu)"
header-includes:
 - \usepackage{amsmath}
 - \usepackage{mathtools}
 - \usepackage{amsthm}
 - \usepackage{multirow}
 - \usepackage{booktabs}
 - \usepackage{minted}
 - \usepackage{color}
 - \usepackage{xcolor}
 - \usepackage{tcolorbox} 
output:
  pdf_document:
    toc: false
    latex_engine: xelatex
    df_print: kable
  html_document:
    df_print: paged
---

\newsavebox{\selvestebox}
\newenvironment{colbox}[1]
  {\newcommand\colboxcolor{#1}%
   \begin{lrbox}{\selvestebox}%
   \begin{minipage}{\dimexpr\columnwidth-2\fboxsep\relax}}
  {\end{minipage}\end{lrbox}%
   \begin{center}
   \colorbox[HTML]{\colboxcolor}{\usebox{\selvestebox}}
   \end{center}}

\newcommand{\var}{\operatorname{Var}}
\newcommand{\expect}{\operatorname{E}}
\newcommand{\corr}{\operatorname{Corr}}
\newcommand{\cov}{\operatorname{Cov}}
\newcommand{\ssr}{\operatorname{SSR}}
\newcommand{\se}{\operatorname{SE}}
\newcommand{\mat}[1]{\mathbf{#1}}
\newcommand{\RR}{\operatorname{RR}}
\newcommand{\eval}[2]{\left. #1 \right\vert_{#2}}

```{r, include=FALSE}
options(tinytex.engine_args = '-shell-escape')
```

Refer to the data from Exercise 1.27

A person's muscle mass is expected to decrease with age.
To explore this relationship in women, a nutritionist randomly selected $15$
women for each $10$ year age group, beginning with $40$ and ending with age $79$.
The input variable $x$ is age (in years), and the response variable $y$ is
muscle mass (in muscle mass units).

# Part (1)
\fbox{Compute the $t$-statistic and $p$-value for testing $H_0 : \beta_1 = 0$.}

We assume a random sample $\{\mathcal{D}_i\}$ where $\mathcal{D}_i = (X_i,Y_i)$ 
is generated by the linear model $Y_i = \beta_0 + \beta_1 X_i + \epsilon_i$. We are
not particularly interested in the distribution of $X_i$ and instead seek to
use it as an explanatory variable (input) for response variable $Y_i$ (output).

Thus, we are interested in estimating the parameters of the conditional
distribution
$$
  Y_i | X_i = x_i \sim \mathcal{N}(\beta_0 + \beta_1 x_i, \sigma^2).
$$

We fit the given data $\{\mathcal{d}_i\}$ where $\mathcal{d}_i = (x_i,y_i)$ to
the linear regression model
$$
  Y_i = \beta_0 + \beta_1 x_i + \epsilon_i
$$
using the following R code:
```{r}
mass.data = read.table('CH01PR27.txt')
colnames(mass.data)=c("mass","age")
mass.mod = lm(mass ~ age, data=mass.data)
```

We compute the $t$-statistic $t^* = \frac{b_1}{\se(b_1)} \sim t(n-2)$ and the
$p$-value $p = \Pr\{t(n-2) > |t_*|\}$ with the following R code:
```{r}
beta1.stats = summary(mass.mod)$coefficients[2,]
beta1.stats
```

We see that $t^* = `r round(beta1.stats[3],digits=3)`$ which has a $p$-value
less than $0.001$.

# Part (2)
\fbox{Provide an interpretation of the above as a test for model compatibility.}

We think of the the null hypothesis as a simplification to the model, which
we \emph{a priori} prefer to the more complicated model (there is a probabilistic
justification for this bias),
$$
  H_0 \colon \beta_1 = 0 \qquad \text{(age has no effect on muscle mass)}.
$$

We obtained a $p$-value of approximately $.000$, and so the data is not
compatible with the no effect model; we accept the model which includes age as a
predictor for muscle mass.

## Alterantive formulation

Can we say the null hypothesis $H_0 : \beta_1 = 0$ is equivalent to
$$
  H_0 : \text{no association between age and muscle mass}
$$
since if $\beta_1 = 0$ then
$$
  Y_1,\ldots,Y_n \overset{\rm{iid}}{\sim} \mathcal{N}(\beta_0,\sigma^2)?
$$

However, I am concerned that I am going beyond the model since our model
is given by the conditional distribution of $Y_i|X_i$. So, we may say that
$x_i$ has no effect on $Y_i$, but in general $y_i$ may effect $X_i$.
Clearly, in this case, we have scientific insight that one's muscle mass does
not effect one's age.

Also, I'm struggling with the wording \emph{compatibility}. I find myself wanting to
say that any outcome that has non-zero probability under the model is compatible
with the model.
Is it simply too subtle to say, "Assuming the null hypothesis is true, then
with probability $p$ the observed statistic or a more extreme value will have
been observed."
This may in fact be a bit much, especially for the less mathematically inclined,
whom we wish to inform with our analysis. So, perhaps we are better off using
phrasing like "fail to reject the null hypothesis" or "the data is not compatible
with the null model." I must confess that I sort of prefer "fail to reject
the null hypothesis," since it seems to be taking the stance that either model
may be correct, but we lack sufficient data to reject the null model, which is
some model we prefer for some reason, e.g., the null model is often the simplest
model in a family of models.

I'm still unsatisfied with my thoughts on this subject. On a related note, I
enjoyed your discussion last Spring on the Bayesian perspective. When we combine
this perspective with, say, a universal prior, perhaps we are getting closer to
a sort of formalism for the scientific method (where the universal prior is a
sort of formalism for Occam's razor).

# Part (3)
\fbox{Provide an interpretation of the above as a test for sign / direction of
the effect.}

Since both $t^* < 0$ and $b_1 < 0$, the data supports the hypothesis of a
negative association between age and muscle mass, i.e., as age increases muscle
mass is expected to decrease.

# Part (4)
\fbox{Compute an interval estimate for $\beta_1$.}

We compute a $95\%$ confidence interval for $\beta_1$ with:
```{r}
alpha = 0.05
beta1.ci = round(confint(mass.mod, level=1-alpha)[2,],digits=3)
beta1.ci
```

We see that $\beta_1$ has a $95\%$ confidence interval given by
$$
  [`r beta1.ci[1]`,`r beta1.ci[2]`].
$$

# Part (5)
\fbox{Explain how an interval estimate provides additional information beyond the test result.}

We can say more about the size of the effect and a measure of the estimation
accuracy.

We make two observations about the CI for $\beta_1$:

1. All the values in the CI are distant from $0$, and thus we can say there is
   strong evidence that age has a large effect.
   
2. The CI represents a relatively small range of values, and thus we can say
   the model is accurate, i.e., the data has a significant amount of information
   about $\beta_1$.
   
# Part (6)
\fbox{Compute an interval estimate for the centered model parameter $\beta_0^*$.}

$\beta_0^*$ represents the mean muscle mass at the mean age in the
sample,
$$
  E(Y|x=\bar{x}) = \beta_0^*,
$$
and thus
$$
  b_0^* = \frac{1}{n} \sum_{i=1}^{n} Y_i.
$$

To compute the confidence interval for $b_0^*$, I refit the model using the
following R code:
```{r}
x = mass.data$age
y = mass.data$mass
x.bar = mean(x)
y.bar = mean(y)
x.star = x - x.bar
star.mod = lm(y ~ x.star, data=mass.data)
beta0.star.ci = round(confint(star.mod)[1,],digits=3)
beta0.star.ci
```

We see that $\beta_0^*$ has a $95\%$ confidence interval given by
$$
  [`r beta0.star.ci[1]`,`r beta0.star.ci[2]`].
$$


# Part (7)
\fbox{Provide an interpretation for your result, stated in the context of the problem.}

We estimate that the mean muscle mass for all women at the center value of
age ($\bar{x} = `r round(x.bar,digits=3)`$) is between `r beta0.star.ci[1]` and
`r beta0.star.ci[2]`.

