---
title: "Test"
author: "Test"
date: "8/5/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Inverse transform method for sampling from random variable with kernel $g$
First, we can use the inverse-transform method.
The kernel of $g$ is $\exp(-|x|)$, and so the normalizing constant $Z$ is given by
$$
  \frac{1}{Z} \int_{-\infty}^{\infty} g(x) dx = 1,
$$
which may be rewritten as
$$
  \int_{-\infty}^{0} e^{x} dx + \int_{0}^{\infty} e^{-x} dx = Z,
$$
which has the solution $Z=2$.
Thus, random variable $Y$ that has a kernel $g$ has a density
$$
  f_Y(y) = \frac{1}{2} \exp(-|x|).
$$

```{r}
g.density <- function(x) { 0.5*exp(-abs(x)) }
```

The cdf $F_Y$ is defined as
$$
  F_Y(y) = \frac{1}{2}\int_{-\infty}^{y} \exp(-|s|) ds.
$$
If $y \leq 0$, then
$$
  F_Y(y) = \frac{1}{2} \int_{-\infty}^{y} \exp(s) ds = \frac{1}{2}\exp(y)
$$
and if $y > 0$, then
$$
  F_Y(y) = F_Y(0) + \frac{1}{2} \int_{0}^{y} \exp(-s) ds = \frac{1}{2} + \frac{1}{2}(1 - \exp(-y)),
$$
which may be written as the piece-wise function
$$
  F_Y(y) =
  \begin{cases}
    \frac{1}{2} \exp(y)                & y \leq 0\\
    1 - \frac{1}{2} \exp(-y) & y > 0.
  \end{cases}
$$

Now, we find the inverse of $F_Y$.
Over $y \leq 0$, the inverse of $F_Y$ is given by
$$
  p = \frac{1}{2} \exp(y).
$$
Solving for $y$, we obtain $y = \log(2p)$. Note that given $y \leq 0$, $p \in (0,0.5]$.
Over $y > 0$, the inverse of $F_Y$ is given by
$$
  p = 1 - \frac{1}{2} \exp(-y).
$$
Solving for $y$, we obtain
$$
  y = -\log(2-2p)
$$
where $p \in (0.5,1)$. We thus have a piecewise function
$$
  F^{-1}(p) =
  \begin{cases}
    \log(2p)        & p \in (0,0.5]\\
    -\log(2-2p)     & p \in (0.5,1).
  \end{cases}
$$
Thus, to sample from $Y$, we observe $u$ from $U(0,1)$ and, if $u \leq 0.5$,
we let $y = \log(2u)$ and otherwise let $y = -\log(2-2u)$.

```{r}
ry <- function(n)
{
  ys <- numeric(n)
  for (i in 1:n)
  {
    u <- runif(1)
    if (u <= 0.5) { ys[i] <- log(2*u) }
    else { ys[i] <- -log(2-2*u)}
  }
  ys
}
```

