---
title: 'Time Series Analysis - 478 - HW #4'
author: "Alex Towell (atowell@siue.edu)"
output:
  pdf_document:
    df_print: kable
    #toc: true
#    toc_depth: 3
    latex_engine: pdflatex
header-includes:
 - \usepackage{amsmath}
 - \usepackage{mathtools}
 - \usepackage{amsthm}
 # - \usepackage{amsbsy}
 - \usepackage{bm}
 - \usepackage{minted}
 - \usepackage{xcolor}
---
\newcommand{\var}{\operatorname{Var}}
\newcommand{\expect}{\operatorname{E}}
\newcommand{\corr}{\operatorname{Corr}}
\newcommand{\cov}{\operatorname{Cov}}
\newcommand{\ssr}{\operatorname{SSR}}
\newcommand{\se}{\operatorname{SE}}
\newcommand{\mat}[1]{\bm{#1}}
\newcommand{\eval}[2]{\left. #1 \right\vert_{#2}}

```{r, include=FALSE}
options(tinytex.engine_args = '-shell-escape')

# moving average 
MA <- function(tsdata, span)
{
   ma=filter(tsdata, rep(1/span,span),side=1)
   return(ma)
}

# exponential moving average
EMA <- function(tsdata, start, discount)
{
   N=length(tsdata)
   ema=vector(length=N)
   ema[1]=start
   for (i in 2:N)
   {
 	   ema[i]=ema[i-1]*(1-discount)+tsdata[i]*discount
   }
   return(ema)	
}

# 2nd-order exponential smoothing
EMA2 <- function(tsdata, start1, start2, discount1,discount2)
{
   ema1=EMA(tsdata, start1, discount1)
   ema2=EMA(ema1, start2, discount2)
   return(list(ema1=ema1,ema2=ema2,yhat=2*ema1-ema2))	
}
```

# Problem 1
\fbox{
\begin{minipage}{.9\textwidth}
Suppose that simple exponential smoothing is being used to forecast a process, $y_t=\mu+\epsilon_t$.
However, at the start of period $t^*$, the mean of the process shifts to a new
mean level $\mu+\delta$.
The mean remains at this new level for subsequent time periods.
Calculate the expected value of the simple exponential moving average.
(Note: you need to discuss cases for $T > t^*$ and $T < t^*$.)
\end{minipage}}

In the simple exponential moving average, we model the mean of the time series with
$$
   \hat{y}_t = \frac{1-\theta}{1-\theta^t} \sum_{j=0}^{t-1} \theta^j y_{t-j}.
$$

The expected value is
$$
   \expect(\hat{y}_t) = \frac{1-\theta}{1-\theta^t} \sum_{j=0}^{t-1} \theta^j \expect(y_{t-j}).
$$
Given $T < t^*$, the process is $y_T = \mu + \epsilon_T$ and given
$T \geq t^*$, the process is $y_T = \mu + \delta + \epsilon_T$.

Letting $T = t-j$ and solving for $j$ in $t-j < t^*$, we see that we may
partition the summation over $j \geq t - t^*$ and $j < t - t^*$,
\begin{align*}
   \expect(\hat{y}_t)
      &= \frac{1-\theta}{1-\theta^t}
      \left\{
         \sum_{j=0}^{t-t^*-1} \theta^j (\mu + \delta)+ \sum_{j=t-t^*}^{t-1} \theta^j \mu
      \right\}\\
      &= \frac{1-\theta}{1-\theta^t}
      \left\{
         \sum_{j=0}^{t-t^*-1} \theta^j \delta + \sum_{j=0}^{t-1} \theta^j \mu
      \right\}.
\end{align*}

Expanding the product,
\begin{align*}
   \expect(\hat{y}_t)
      &= \frac{1-\theta}{1-\theta^t}\delta \sum_{j=0}^{t-t^*-1} \theta^j  + \mu\\
      &= \frac{1-\theta}{1-\theta^t}\delta \frac{1-\theta^{t-t^*}}{1-\theta}  + \mu\\
      &= \frac{1-\theta^{t-t^*}}{1-\theta^t}\delta + \mu\\
      &= \left(
         \frac{1}{1-\theta^t} -
         \frac{\theta^{t-t^*}}{1-\theta^t}
         \right) \delta + \mu.
\end{align*}

For large $t$, $\theta^t \approx 0$, and so we simplify the above to
$$
   \expect(\hat{y}_t) =
   \left(
      1 - \theta^{t-t^*}
   \right) \delta + \mu.
$$
We consider two degenerate cases given by $t^* = t$ and $t^* = 0$
which respectively have expectations given by $\expect(\hat{y}_t) = \mu$
and $\expect(\hat{y}_t) = \sigma + \mu$ (when we use the assumption that
$\theta_t \approx 0$).




# Problem 2

Consider the time series $\{y_T\} = (14,19,18,22,17,28,43,45,62,60)$.

## Parts (a)-(d)

We use the following R code to generate and populate the table.
```{r}
discount <- 0.2
tsdata <- ts(c(14,19,18,22,17,28,43,45,62,60))

# part (a): calculate the simple moving average with span=3.
ma3 <- MA(tsdata,3)

# part (b): calculate the simple (1st order) exponential moving average using
# discount lambda = 0.2 and initial value y1.
#ema1 <- EMA(tsdata,tsdata[1],discount)
ema1 <- EMA(tsdata,14,discount)

# part (c): calculate the 2nd order exponential moving average using discount
# lambda = 0.2 and initial values y1.
ema2_data <- EMA2(tsdata,tsdata[1],tsdata[1],discount,discount)
#ema2_data <- EMA2(tsdata,tsdata[1],tsdata[1],discount,discount)
ema2 <- as.numeric(unlist(ema2_data[2]))

# part (d): calculate the (unbiased) linear trend estimator
yhat <- as.numeric(unlist(ema2_data[3]))

prob2 <- data.frame(
   t = 1:length(tsdata),
   yt = tsdata,
   ma = ma3,
   ema1 = ema1,
   ema2 = ema2,
   yhat = yhat,
   delta = tsdata-yhat)

# parts (a)-(d): generate the table with the previously computed values
knitr::kable(prob2,caption="Parts (a)-(d)",padding=-1L,
             col.names =c("time",
                          "$y_T$",
                          "MA","$\\hat{y}_T^{(1)}$",
                          "$\\hat{y}_T^{(2)}$",
                          "$\\hat{y}_T$",
                          "$y_T-\\hat{y}_T$"))
```

## Part (e)
\fbox{Finish the table by calculate the errors. What is the SSE?}
The sum of squared error (SSE) is computed and displayed by the following
R code:

```{r}
sse = sum((tsdata-yhat)^2)
cat("The sum of squared error is ", sse, ".")
```

Note that this is not the forecast one-step sum of squared error.
I believe you would specifically ask for that if that is what you wanted,
and the computation of the one-step ahead SSE seemed a bit tedious.

## Part (g)
\fbox{\begin{minipage}{.8\textwidth}
Make a one-step-ahead forecast for Time=11 based on the linear trend process.
If the true observation is 72, what’s your prediction error?
Also provide the prediction interval.
\end{minipage}}

```{r message=FALSE}
# here's how we can compute the one-step-ahead forecast.
# i decided to do this computation manually, rather than using the forecasting
# library.
b1hat <- (ema1[10] - ema2[10]) * discount / (1-discount)
y10_one_step <- yhat[10] + b1hat
print(y10_one_step)
```

We see that $\hat{y}_{11}(10) \approx 57$.
If the true observation $y_{11} = 72$, then the $1$-step forecast error is
$$
   e_{10}(1) = y_{11} - \hat{y}_{11}(10) = 72 - 57 = 15.
$$

To compute the intervals, we use the forecast library.
```{r message=FALSE}
library("forecast")
dEMA <- holt(tsdata,h=1,level=c(95),initial="simple",alpha=discount,beta=discount)
summary(dEMA)

# we got this estimate of one-step-ahead variance of
# the forecast errors from the holt procedure
rmse <- 9.238
lo <- y10_one_step - 1.96 * rmse
hi <- y10_one_step + 1.96 * rmse
print(y10_one_step)
print(lo)
print(hi)
```

The $95\%$ prediction interval computed from the \emph{forecast} library is
$$
   [40.78, 76.99].
$$

As a slight twist, I thought I would use the estimate of the variance of
the forecast errors from the \emph{holt} procedure's output and apply it
to the previous approach, resulting in the interval
$$
   [38.94, 75.16].
$$

Observe that these prediction interval estimates differ, primarily due to the
fact that the \emph{holt} procedure uses a different set of initial values,
but they are reasonably close.

# Problem 3
Consider the Dow Jones Index data on Blackboard.
The dataset contains yearly Dow Jones closing index from year 1981 to 2016.

## Part (a)
\fbox{Read the data into R. Then construct a time plot ($D_j$ index v.s. time).}

The following R code is used to read the data file and generate the time series plot:
```{r}
library(latex2exp)
data <- read.table("./DJI_yearly.txt", header=TRUE)
tsdata <- ts(data[,2])
plot(tsdata, main=TeX("$D_j$ index versus time"), xlab="time", ylab=TeX("$D_j$ index"))
```

## Part (b)
\fbox{\begin{minipage}{0.8\textwidth}
Calculate the simple exponential moving average of the data using $\lambda = 0.1$,
and initial value equals the 1st observation.
Impose the 1st order EMA on the time plot.
Comment on what you observe.
\end{minipage}}

```{r}
discount <- 0.1
ema1 <- EMA(tsdata,tsdata[1],discount)
plot(tsdata,type="b",col="blue",pch='*',ylab=TeX('$\\hat{y}$'), main=TeX('$y$ versus time'))
lines(ema1,type="l",col="orange")
legend("topleft", 
  legend = c("Data", TeX("$\\hat{y}_T^{(1)}$")), 
  col = c(
     "blue", 
     "orange"), 
  pch = c('*','-'))
```

The first-order EMA $\hat{y}_T^{(1)}$ has similiar growth to the time series
data, but it consistently estimates values below it.

## Part (c)
\fbox{Calculate the sum of squared error (SSE) of your 1st order EMA.}

The following R code is used to compute and display the SSE.
```{r}
sse = sum((tsdata-ema1)^2)
cat("The sum of squared error is ", sse, ".")
```

## Part (d)
\fbox{\begin{minipage}{.8\textwidth}
Make a one-step-ahead forecast for year 2017 based on 1st order EMA.
Also provide the prediction interval.
\end{minipage}}

```{r warning=FALSE, message=FALSE}
library("forecast")
sEMA=ses(tsdata, h=1, level=c(95), initial = "simple", alpha = discount)
summary(sEMA)
```

First, note that the year $2017$ corresponds to time unit $37$ and the $1$-step
ahead forecast for $2018$ is therefore time unit $38$.
Thus, we see that $\hat{y}_{38}^{(1)}(37) = 13733.58$.
The $95\%$ prediction interval is $[5046.53,22420.63]$.

## Part (e)
\fbox{
\begin{minipage}{.8\textwidth}
Fit a linear regression $y_t = \beta_0 + \beta_1 t + \epsilon_t$ with the data
and report the OLS estimators $\hat\beta_0$ and $\hat\beta_1$.
\end{minipage}}

We denote these estimators by $\hat\beta_{0,0}$ and $\hat\beta_{1,0}$, since
they are from OLS given the data and the model $Y_t = \beta_0 + \beta_1 t + \epsilon_t$.

```{r}
t <- 1:length(tsdata)
ols <- lm(tsdata~t) 
b00=ols$coeff[1]
b10=ols$coeff[2]
print(ols)
```
We see that $\hat\beta_{0,0} = -1620.7$ and $\hat\beta_{1,0} = 527.4$.

## Part (f)
\fbox{Set the initial values for 2nd order smoothing based on part (e).}

```{r}
y01=b00-(1-discount)/discount*b10
y02=b00-2*(1-discount)/discount*b10
cat("y01 =", y01, "and y02 =", y02, ".")
```

We see that the initial values are given by $\hat{y}_0^{(1)} = -6367.144$ and
$\hat{y}_0^{(2)} = -11113.57$.

## Part (g)
\fbox{\begin{minipage}{.8\textwidth}
Calculate the 2nd order exponential moving average of the data using $\lambda=0.1$, and use initial
values in part (f).
Calculate the unbiased estimates, by $\hat{y}_T = 2\hat{y}_T^{(1)} - \hat{y}_T^{(2)}$.
Plot the estimates on the original time plot.
Compare with the 1st order EMA approach.
Comment on what you observe.
\end{minipage}}

```{r warning=FALSE, message=FALSE}
ema2 <- EMA2(tsdata,y01,y02,discount,discount)

plot(tsdata,col="black",type="b",pch='*')
lines(ema2$ema1,col="red")
lines(ema2$ema2,col="blue")
lines(ema2$yhat,col="green")

legend("topleft",
  legend = c(
     "data",
     TeX("$\\hat{y}_T^{(1)}$"),
     TeX("$\\hat{y}_T^{(2)}$"),
     TeX("$\\hat{y}_T$")),
  col = c(
     "black",
     "red",
     "blue",
     "green"),
  pch = c('*','-','-','-'))
```

We plot the data, a realization of $\{Y_t\}$, in black, $\hat{y}_T^{(1)}$ in
red, $\hat{y}_T^{(2)}$ in blue, and $\hat{y}_T = 2 \hat{y}_T^{(1)} - \hat{y}_T^{(2)}$
in green.

Each of the estimators exhibits a similiar trend. However, the most biased
is the second-order smoother $\hat{y}_T^{(2)}$, $\hat{y}_T^{(1)}$ exhibits less
bias, and $\hat{y}_T$ seems relatively unbiased, although the original time series
may have a higher order component.

## Part (h)
\fbox{Calculate the sum of squared error (SSE) of your estimators in part (g).}

```{r}
print(sum((tsdata-ema2$ema1)^2))
print(sum((tsdata-ema2$ema2)^2))
print(sum((tsdata-ema2$yhat)^2))
```

Note that I am assuming we are using the normal sum of squared errors, e.g.,
$\operatorname{SSE} = \sum_{t=1}^{T} (y_t - \hat{y}_t)$.

We see that the SSE for $\hat{y}_T^{(1)}$ is $1063748664$, the SSE for
$\hat{y}_T^{(2)}$ is $3721105725$, and the SSE for $\hat{y}_T$ is $103638848$.

## Part (i)
\fbox{\begin{minipage}{.8\textwidth}
Make a one-step-ahead forecast for year 2017 based on 2nd order EMA.
Also provide the prediction interval.
\end{minipage}}

I'm going to use the \emph{forecast} library, which will also compute a
prediction interval.

```{r}
step1=ses(tsdata, h=1, level=c(95), initial="simple", alpha=discount)
step2=ses(step1$fitted, h=1, level=c(95), initial="simple", alpha=discount)
summary(step2)
```

We see that $\hat{y}_{38}^{(2)}(37) \approx 8311$ with a $95\%$ prediction
interval $[3356.25, 13264.82]$.

Note that we get much better results if we use the \emph{holt} procedure,
which corrects for the bias.
```{r}
improved=holt(tsdata, h=1, level=c(95), initial="simple", alpha=discount, beta=discount)
summary(improved)
```
We see that $\hat{y}_{38}(37) \approx 18370$ with a $95\%$ prediction
interval $[13558.89, 23180.81]$.

## Part (j)
\fbox{\begin{minipage}{.8\textwidth}
The true index for 2017 is 24719.22.
What is your forecast errors in part (d) and (i).
Is the 2nd order approach an apparent improvement over the use of simple
exponential smoothing?
\end{minipage}}

The forecast error for the simple exponential smoother $\hat{y}_38^{(1)}(37)$
is
$$
   e_37(1) = 24719.22 -13733.58 \approx 10986
$$
and the forecast error for the double exponential smoother $\hat{y}_38^{(2)}(37)$
is
$$
   e_37(1) = 24719.22 - 8311 \approx 16408.
$$
The second-order smoother has significantly more forecast error.

If we use the results from the \emph{holt} procedure, which corrects for the bias,
we get a much better result,
$$
   e_37(1) = 24719.22 - 18370 \approx 6349.
$$

# Appendix
\begin{figure}[h]
\centering
\begin{minted}{R}
# moving average 
MA <- function(tsdata, span)
{
   ma=filter(tsdata, rep(1/span,span),side=1)
   return(ma)
}

# exponential moving average
EMA <- function(tsdata, start, discount)
{
   N=length(tsdata)
   ema=vector(length=N)
   ema[1]=start
   for (i in 2:N)
   {
 	   ema[i]=ema[i-1]*(1-discount)+tsdata[i]*discount
   }
   return(ema)	
}

# 2nd-order exponential smoothing
EMA2 <- function(tsdata, start1, start2, discount1,discount2)
{
   ema1=EMA(tsdata, start1, discount1)
   ema2=EMA(ema1, start2, discount2)
   return(list(ema1=ema1,ema2=ema2,yhat=2*ema1-ema2))	
}
\end{minted}
\caption{Library of functions}
\end{figure}