---
title: 'STAT 581 - HW #1'
author: "Alex Towell (atowell@siue.edu)"
output:
  #md_document:
  #  variant: markdown_github
  pdf_document:
    df_print: kable
    latex_engine: xelatex
    #keep_tex: true
    highlight: tango
    #highlight: breezedark
    #highlight: pygments
    #highlight: zenburn
    #highlight: haddock
    #highlight: kate
    #highlight: espresso
fontsize: 11pt
geometry: margin=.5in
documentclass: article
#documentclass: paper
#documentclass: standalone
#classoption: twocolumn
header-includes:
 - \usepackage{amsmath}
 - \usepackage{mathtools}
 - \usepackage{amsthm}
 # - \usepackage{amsbsy}
 - \usepackage{bm}
 - \usepackage{xcolor}
 - \usepackage{fbox}
 - \usepackage{amsmath}
 - \usepackage{hw4}
---

```{r echo=F}
#knitr::opts_chunk$set(tidy.opts = list(width.cutoff = 80), tidy = TRUE)
knitr::opts_chunk$set(tidy.opts = list(width.cutoff = 80))
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```


```{r}
library("readxl")

# read from directory the markdown file is located in
h1.data <- read_excel("./handout1data.xlsx")
h1.data$method <- as.factor(h1.data$method)

h1.time.dry <- na.omit(h1.data$time[h1.data$method=='d'])
h1.time.wet <- na.omit(h1.data$time[h1.data$method!='d'])
```



An experiment is designed to investigate whether the time to drill holes in rock
holes using wet or dry drilling.
A completely randomized design (CRD) is used.
Each method is replicated on 12 rocks.
The drilling times (in 1 / 100 minutes) are observed to be
```{r echo=F}
library(printr)
t(data.frame(dry=h1.time.dry,wet=h1.time.wet))
```

# Q1
\fbox{State the hypotheses of interest. Provide an interpretation, stated in the
context of the problem.}

The hypothesis of interest is whether drilling time is effected by the choice
of wet or dry drilling. We may formulate this as a hypothesis test of the form

\begin{align*}
  H_0 &: \mu_1 = \mu_2\\
  H_A &: \mu_1 \neq \mu_2\,,
\end{align*}
where $\mu_1$ and $\mu_2$ are respectively the expected drilling time for
dry and wet drilling.

If $H_0$ is true, the drilling method has \emph{no} effect on drilling time. If
$H_A$ is true, the choice does have an effect on drilling time.

# Q2
\fbox{Give two advantages of the completely randomized design.}

Two advantages of the completely randomized design (CRD) are:

  1. It is simple to implement and analyze.
  
  2. It accounts for experimental unit variance, namely by controlling for the
     type I error,
     
     $$
         \alpha = \Pr(\text{decide} \, H_A | H_0 \, \text{true}),
     $$
     
     where $\alpha$ is the probability of a type I error. It may also be thought of
     as the type I \emph{error rate}, or \emph{false positive rate}.

# Q3
\fbox{Give two disadvantages of the completely randomized design.}

Two disadvantages of the completely randomized design (CRD) are:

  1. It does not account for type II error,
     $$
         \beta = \Pr(\text{decide} \, H_0 | H_A \, \text{true}).
     $$
     where $\beta$ is the probability of a type II error.

     > **Additional remarks**: Since we may denote a type I error a false
     positive, it may be appropriate to denote a type II error a false negative,
     which implies the convention $H_0 \equiv \rm{negative}$ and
     $H_A \equiv \rm{positive}$. Thus, $\beta$ may be thought of as the false
     negative rate.

     > Consider the unrelated *Bloom filter*, which is a
     probabilistic data structure that models sets in which membership queries
     have a controllable false positive rate $\epsilon$ and a false negative rate
     $0$. If we take the complement of an object representing such a Bloom
     filter, then membership queries on its complement have a false positive
     rate $0$ and a false negative rate $\epsilon$.
     
  2. It does not adjust for differences in experimental units nor differences
  in any other factors.
  

# Q4
\fbox{Compute the sample means, the sample standard deviations, and the pooled
sample standard deviation.}

```{r}
time.mean <- round(c(mean(h1.time.dry),
               mean(h1.time.wet)),digits=3)
time.var <- round(c(var(h1.time.dry),
              var(h1.time.wet)),digits=3)
time.sd <- round(sqrt(time.var),digits=3)
n <- c(length(h1.time.dry),
       length(h1.time.wet))
time.sd.pooled <-
  round(sqrt(((n[1] - 1) * time.var[1] + (n[2] - 1) * time.var[2]) / (sum(n)-2)),
        digits=3)

c("y.bar.1" = time.mean[1],
  "y.bar.2" = time.mean[2],
  "s.1" = time.sd[1],
  "s.2" = time.sd[2],
  "s.p" = time.sd.pooled)
```

We see that the sample means are
$$
  (\bar{y}_1,\bar{y}_2) = (`r time.mean`),
$$
the standard deviations are
$$
  (s_1,s_2) = (`r time.sd`),
$$
and the pooled standard deviation is
$$
  s_p = `r time.sd.pooled`.
$$

# Q5
\fbox{Compute the $t_0$ statistic, the critical point for a level
$\alpha = .05$ test, and the $p$-value.}

The test statistic is defined as
$$
  t_0 = \frac{\bar{y}_1-\bar{y}_2}{s_p \sqrt{\frac{1}{n_1}+\frac{1}{n_2}}}.
$$
which is a drawn from a $t$-student distribution with $n_1+n_2-2$ degrees of
freedom under the null model (reference distribution).

> Under the null model, $\bar{y}_1$ and $\bar{y}_2$ are the means of
two samples respectively of size $n_1$ and $n_2$ whose elements are drawn from
the same normal distribution.

Denote this reference distribution by $T$.
Under the null model, $t_0$ is a realization of $T$.
Our general approach is to compute the test statistic $t_0$ for the given
samples and say it is compatible with the null model if it obtains a value that
is not too extreme under $T$, e.g., a $p$-value that is less than some
controllable parameter $\alpha$.

The critical point (the minimum value that is considered to be too extreme) is
given by solving for $t^*$ in
$$
  \Pr\{-|t^*| < T < |t^*|\} = \alpha,
$$
which has the solution $t^* = t_{\rm{df}=n_1+n_2-2}(\alpha/2)$, and
\begin{align*}
  \text{$p$-value} &= \Pr\{T \notin [-|t_0|,|t_0|]\}\\
                   &= 2 \Pr\{T > |t_0|\}\\
                   &= 2(1-F_T(|t_0|)),
\end{align*}
where $F_T$ is the cdf of $T$.

We compute these results in R with:
```{r}
se <- time.sd.pooled*sqrt(1/n[1]+1/n[2])
t.0 <- round((time.mean[1]-time.mean[2])/se,digits=3)
alpha <- 0.05
t.star = round(qt(alpha/2,lower.tail=F, df=n[1]+n[2]-2),digits=3)
p.value = 2*pt(abs(t.0),df=n[1]+n[2]-2,lower.tail=F)

c("t.0"=t.0,"t.*"=t.star,"p.value"=p.value)
```

We see that $t_0 = `r t.0`$, the critical point $t^*$ for $\alpha = 0.05$ is $`r t.star`$,
and the $p$-value is less than $.001$.

<!-- An alternative way to compute these results is provided by -->
<!-- $\operatorname{t.test}$: -->
<!-- ```{r} -->
<!-- method <- as.factor(na.omit(h1.data$method)) -->
<!-- times <- na.omit(h1.data$time) -->
<!-- t.test(formula=times~method,       # two-sample t.test -->
<!--        mu=0,                       # mu denotes difference in means: mu.1-mu.2=0 -->
<!--        alternative=c("two.sided"), # H.0 : mu.1=mu.2, H.A : mu.1 != mu.2 -->
<!--        var.equal=T)                # treat the variances as being equal -->
<!-- ``` -->

# Q6
\fbox{Provide an interpretation, stated in the context of the problem.}

The experiment finds that the wet drilling method leads to reduced drilling
times compared to the dry drilling method.

# Q7
\fbox{Create a boxplot as a graphical display of the data.}

```{r}
method <- na.omit(h1.data$method)
times <- na.omit(h1.data$time)
boxplot(times~method,names=c("dry","wet"),xlab="method",ylab="time")
```

# Q8
\fbox{\begin{minipage}{\textwidth}
Refering to the boxplot, is it true that all dry drilling times exceed all wet
drilling times?
In what sense can the experiment find that dry drilling takes longer than wet
drilling?
\end{minipage}}

No, not all dry drilling times exceed all wet drilling times.

However, the boxplot does show that the spread of the drilling times area
approximately the same between the two methods (providing justification for
assuming the two samples have the same variance) and the mean drilling times are
significantly different. As a result, most of the time, the wet drilling time is
going to be less than the dry drilling time.

For instance, we see that the mean wet drilling time is less than the best dry
drilling time and we also see that the $25$-th percentile of the dry method's
times is a bit larger than the wet method's longest drilling time.
Most tellingly, we see that around $67\%$ of the time, the wet drilling time
is less than the smallest dry drilling time.