---
title: 'STAT 581 - Problem Set 6'
author: "Alex Towell (atowell@siue.edu)"
output:
  #md_document:
  #  variant: markdown_github
  pdf_document:
    df_print: kable
    latex_engine: xelatex
    #keep_tex: true
    highlight: tango
    #highlight: breezedark
    #highlight: pygments
    #highlight: zenburn
    #highlight: haddock
    #highlight: kate
    #highlight: espresso
fontsize: 11pt
geometry: margin=.5in
documentclass: article
#documentclass: paper
#documentclass: standalone
#classoption: twocolumn
header-includes:
 - \usepackage{amsmath}
 - \usepackage{mathtools}
 - \usepackage{amsthm}
 # - \usepackage{amsbsy}
 - \usepackage{bm}
 - \usepackage{xcolor}
 - \usepackage{fbox}
 - \usepackage{amsmath}
 - \usepackage{hw4}
---


```{r echo=F}
#knitr::opts_chunk$set(tidy.opts = list(width.cutoff = 80), tidy = TRUE)
knitr::opts_chunk$set(tidy.opts = list(width.cutoff = 80))
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

# Problem 1

A soft drink bottler is interested in studying the effects on a filling process.
A factorial experiment is run using three factors: percent carbonation (in %),
operating pressure (in psi), and line speed (in bpm).
The response variable is deviation from the target fill height.
The experiment consists of $n=2$ runs for each of the $a b c = 12$ treatment
combinations.
The data is available on Blackboard as an Excel File.

## Preliminary analysis

The three-factor factorial experiment with $a=3$ levels of the *carbonation*
factor (in \%), denoted by $A$, $b=2$ levels of the operating *pressure* (in psi),
denoted by $B$, and $c=2$ levels of the line *speed* (in bpm), denoted by $C$.
Each treatment combination has $n=2$ replicates for a total of $N=a b c n=24$
observations.
The response variable $y$ is deviation from the target fill height.

We describe the observations in the three-factor factorial experiment by the
fixed-effects interaction model where treatments are defined to be deviations from the
overall mean,
$$
  Y_{i j k \ell} = \mu + \tau_i + \beta_j + \gamma_k +
                   (\tau\!\beta)_{i j} + (\tau\!\gamma)_{i k} +
                   (\beta\!\gamma)_{j k} + (\tau\!\beta\gamma)_{i j k} +
                   \epsilon_{i j k \ell}
  \begin{cases}
    i = 1,\ldots,a=3\\
    j = 1,\ldots,b=2\\
    k = 1,\ldots,c=2\\
    \ell = 1,\ldots,n=2\\
  \end{cases}
$$
where $\mu$ is the overall main effect, $\tau_i$ is the effect of the $i$-th
level of factor $A$, $\beta_j$ is the effect of the $j$-th level of factor $B$,
and $\gamma_k$ is the effect of the $k$-th level of factor $C$.
The interaction effects may also be observed using the same labeling, e.g.,
$(\tau\!\gamma)_{i k}$ is the interaction effect of the $i$-th level of factor
$A$ and the $k$-th level of factor $C$.

## (a)
\fbox{\begin{minipage}{\textwidth}
Perform tests for all main effects and for all interaction effects.
State the $F$-statistic and $p$-value for each of the main effect tests, and for
the test of an interaction between carbonation and pressure.
\end{minipage}}

```{r}
library("readxl")
h6.data = read_excel("handout6data.xlsx")

A = as.factor(na.omit(h6.data$carbon))
B = as.factor(na.omit(h6.data$pressure))
C = as.factor(na.omit(h6.data$speed))
y = na.omit(h6.data$deviation)

contrasts(A) = contr.sum # we use these in the additional analysis section
contrasts(B) = contr.sum # to make it easier to provide explicit estimates of
contrasts(C) = contr.sum # the parameters

x.mod = aov(y ~ A*B*C)
summary(x.mod)
```

The tests for each of the main effects and a test of an interaction between
carbon $(A$) and pressure $(B$) is given by the following table:

| effects      | $F^*$   | $p$-value |
|--------------|---------|-----------|
| $A$          | $178.4$ | $.000$    |
| $B$          | $64.1$  | $.000$    |
| $C$          | $31.1$  | $.000$    |
| $A \times B$ | $3.7$   | $.056$    |

## (b)
\fbox{\begin{minipage}{\textwidth}
Explain how an interaction plot is used in studying an interaction effect.
\end{minipage}}

If an interaction plot is parallel, or nearly so, then there is no need
to include the interaction term.

In the $3$-factor ANOVA model, a $3$-factor interaction occurs when a $2$-factor
interaction depends on the level of the $3$rd factor.

## (c)
\fbox{\begin{minipage}{\textwidth}
Determine fitted values for the main effects model, as well as fitted values for
the model including the carbonation / pressure interaction.
Plot the fitted values for each model using an interaction plot.
Comment on the decision of whether or not to include the interaction term.
\end{minipage}}

From the earlier results, we saw that factors $A$ (carbon), $B$ (pressure),
and $C$ (speed) were statistically significant, but only the interaction between
factor $A$ and $B$ (carbon and pressure) had any significance.

### Additive model
The additive model is given by
$$
  Y_{i j k \ell}^{(R)} = \mu + \tau_i + \beta_j + \gamma_k + \epsilon_{i j k \ell}.
$$
Thus,
$$
  E(Y_{i j k}^{(R)}) = \mu + \tau_i + \beta_j + \gamma_k.
$$

We fit the additive model with:
```{r}
mod.R = aov(y ~ A+B+C)
```

We define the estimated means for the models with:
```{R}
fits.R = predict(mod.R)
```

The interaction plots for the additive where the left and right plots are
respectively for factor $C$ at levels $1$ and $2$.
```{r}
par(mfrow=c(1,2))
for (i in 1:nlevels(C))
{
  interaction.plot(A[C==levels(C)[i]],B[C==levels(C)[i]],
                   fits.R[C==levels(C)[i]],
                   xlab="A",
                   ylab="mean response",
                   trace.label="B")
}
```

### Interaction model
The model that includes interactions between factors $A$ (carbon) and $B$
(pressure) is given by
$$
  Y_{i j k \ell}^{(F)} = \mu + \tau_i + \beta_j + \gamma_k +
                   (\tau\!\beta)_{i j} + \epsilon_{i j k \ell}.
$$
Thus,
$$
  E(Y_{i j k}^{(F)}) = \mu + \tau_i + \beta_j + \gamma_k + (\tau\!\beta)_{i j}.
$$

We fit the interaction model with:
```{r}
mod.F = aov(y ~ A+B+C+A:B)
```

We define the estimated means for the models with:
```{R}
fits.F = predict(mod.F)
```

The interaction plots for the interaction model  where the left and right plots
are respectively for factor $C$ at levels $1$ and $2$.
```{r}
par(mfrow=c(1,2))
for (i in 1:nlevels(C))
{
  interaction.plot(A[C==levels(C)[i]],B[C==levels(C)[i]],
                   fits.F[C==levels(C)[i]],
                   xlab="A",
                   ylab="mean response",
                   trace.label="B")
}
```

### Decision
The interaction model fits show only a slight difference difference from the
additive model.
As a general rule, we prefer simpler models to complex models, and so we decide
to leave out the interaction term.

## (d)
\fbox{\begin{minipage}{\textwidth}
Create box plots showing the main effects.
Provide an overall conclusion, stated in the context of the problem.
\end{minipage}}

```{r}
par(mfrow=c(1,3))
plot(fits.R ~ A)
plot(fits.R ~ B)
plot(fits.R ~ C)
```

We see that each factor has a positive effect on fill height.
The factor with the biggest marginal effect on fill height is the percent of
carbonation (factor $A$).

# Problem 2
An engineer is interested in the effects of cutting speed, tool geometry, and
cutting angle on the life (in hours) of a machine tool.
Two levels of each factor are set, and $n=3$ runs of a $2^3$ design are
completed.
The data is available on Blackboard as an Excel File.

## (a)
\fbox{\begin{minipage}{\textwidth}
Perform tests for all main effects and for all interaction effects.
State the $F$-statistic and $p$-value for each test of an effect deemed to be
important.
\end{minipage}}

```{r}
data2 = read_excel("handout6data.xlsx")

A = as.factor(na.omit(data2$spd))
B = as.factor(na.omit(data2$geo))
C = as.factor(na.omit(data2$angle))
y = na.omit(h6.data$life)

contrasts(A) = contr.sum
contrasts(B) = contr.sum
contrasts(C) = contr.sum

saturated.mod = aov(y ~ A*B*C)
summary(saturated.mod)
```

Factors $B$ and $C$ are important effects and the the interaction between $A$
and $C$ is also important.
Thus, $A$, $B$, $C$, and $A:C$ should be included in the model (we include $A$
because of the importance of $A:C$).

| effects | $F^*$    | $p$-value |
|---------|----------|-----------|
| $A$     | $.022$   | $.884$    |
| $B$     | $25.547$ | $.000$    |
| $C$     | $9.287$  | $.008$    |
| $A:C$   | $15.519$ | $.001$    |

## (b)
\fbox{\begin{minipage}{\textwidth}
Fit a reduced model with the main effects, and the statistically significant
interaction effect.
\end{minipage}}

```{r}
reduced.mod = aov(y ~ A+B+C+A:C)
summary(reduced.mod)
```

### (i)
\fbox{\begin{minipage}{\textwidth}
Create a plot for the interaction effect.
Provide an interpretation, stated in the context of the problem.
\end{minipage}}

```{r}
# define the estimated means for the reduced model for plotting (randomness has
# been smoothed)
fits.r = predict(reduced.mod)
interaction.plot(A,C,fits.r,ylab="y")
```

#### Interpretation
The experiment finds that when cutting speed (factor $A$) is low, cutting angle
has a positive effect on the life of a machine tool (response).

### (ii)
\fbox{\begin{minipage}{\textwidth}
Create a box plot showing the main effect for the remaining factor.
Provide an interpretation, stated in the context of the problem.
\end{minipage}}

```{r}
plot(fits.r ~ B)
```

#### Interpetation
Tool geometry (factor $B$) has a positive effect on the life of a machine tool.

### (iii)
\fbox{\begin{minipage}{\textwidth}
Create a plot of the fitted values for the reduced model.
Which setting of the factors should be used if the goal is to maximize tool
life.
Explain how this result follows from the interaction plot and the main effects
plot.
\end{minipage}}

```{r}
A.C = interaction(A,C)
interaction.plot(B,A.C, fits.r)
```

#### Maximum setting
In this characterization experiment, we find that the best combination of
factor levels is given by a low cutting speed ($A^-$), a high tool geometry
($B^+$), and a high cutting angle ($C^+$).

#### Explanation
First, we observe that $A^-$ and $C^+$ has the best interaction effect (its
corresponding line is above all of the others).
Second, we see that $B$ has a positive effect, i.e., as we go from $B^-$ to
$B^+$, the response increases, and thus $B^+$ is better.
Taking these two facts together, we see that the best combination in this
experiment is $(A^-,B^+,C^+)$.

We may also just look at the highest point on the graph and see that
$(A^-,B^+,C^+)$ has the highest mean response.
However, the interaction plot reveals more, particularly if in follow-up
experiments where we use what we learned in this experiment to narrow down
our search for an optimal combination (where optimality may be with respect to,
say, mean response or minimal variability in the response).

# Appendix
## Problem 1: additional analysis of the additive model
The model parameter estimates are given by:
```{r}
dummy.coef(mod.R)
```

Thus, the estimated mean response $\hat{Y}_{i j k}^{(R)}$ is given by
\begin{align*}
  \hat{Y}_{i j k}^{(R)} = 3.125
    - &I(i=1) 3.625 - I(i=2) 6.25 + I(i=3) 4.250\\
    - &I(j=1) 1.375 + I(j=2) 1.375\\
    - &I(k=1) 9.58\bar{3} + I(k=2) 9.58\bar{3}.
\end{align*}

In R code, this may be brute-force implemented as:
```{r}
yhat.R = function(i,j,k)
{
  y = 3.125
  if (i == 1) y = y - 3.625; if (i == 2) y = y - .625; if (i == 3) y = y + 4.25
  if (j == 1) y = y - 1.375; if (j == 2) y = y + 1.375
  if (k == 1) y = y - .95833333; if (k == 2) y = y + .95833333
  y
}
```

We print out all of the estimates of the mean response with:
```{r}
for (i in 1:3)
  for (j in 1:2)
    for (k in 1:2)
      cat("Y(",c(i,j,k),") = ",yhat.R(i,j,k),"\n")
```


## Problem 1: plots of interactions that were deemed insignificant
The data showed no statistically significant interaction effects between
factors $A$ and $B$ and between factors $B$ and $C$.
Thus, their interaction plots are, as expected, parallel:
```{r}
par(mfrow=c(1,2))
interaction.plot(A,B,fits.r,ylab="y")
interaction.plot(B,C,fits.r,ylab="y")
```

