---
title: 'STAT 581 - Problem Set 3.b'
author: "Alex Towell (atowell@siue.edu)"
output:
  #md_document:
  #  variant: markdown_github
  pdf_document:
    df_print: kable
    latex_engine: xelatex
    #keep_tex: true
    highlight: tango
    #highlight: breezedark
    #highlight: pygments
    #highlight: zenburn
    #highlight: haddock
    #highlight: kate
    #highlight: espresso
fontsize: 11pt
geometry: margin=.5in
documentclass: article
#documentclass: paper
#documentclass: standalone
#classoption: twocolumn
header-includes:
 - \usepackage{amsmath}
 - \usepackage{mathtools}
 - \usepackage{amsthm}
 # - \usepackage{amsbsy}
 - \usepackage{bm}
 - \usepackage{xcolor}
 - \usepackage{fbox}
 - \usepackage{amsmath}
 - \usepackage{hw4}
---

```{r echo=F}
#knitr::opts_chunk$set(tidy.opts = list(width.cutoff = 80), tidy = TRUE)
knitr::opts_chunk$set(tidy.opts = list(width.cutoff = 80))
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

# Problem 3
An experiment to compare a new drug to a standard is in the planning stages.
The response variable of interest is the clotting time (in minutes) of blood
drawn from a subject.
The experimenters want to perform a two sample $t$ test at level $\alpha = .05$,
having power $\pi=.8$ at $\delta_A=0.25$, for standard deviation $\sigma=0.7$.

## Part (a)
\fbox{\begin{minipage}{\textwidth}
Determine the sample size for each drug in order to achieve the stated test
specifications.
\end{minipage}}

Selection of sample size is of fundamental importance in experimental design.
We know that the CI for the difference in two means has a length given by
$$
  t_{\alpha/2,n_1+n_2-2} s_p \sqrt{\frac{1}{n_1} + \frac{1}{n_2}}
$$
and if $n_1=n_2=n$ then this simplifies to
$$
  t_{\alpha/2,2n-2} s_p \sqrt{\frac{2}{n}}.
$$

We cannot control $s_p$ but we can control $n$.
While $s_p$ will be relatively constant, $\sqrt{n^{-1}}$ will decrease as $n$
increases, so the length of the CI is approximately $\mathcal{O}(\sqrt{n^{-1}})$.

A small CI is desirable, but in particular, in hypothesis testing, if $H_A$ is
true we may be concerned about type II errors, whose probability $\beta$ is not
only a function of $\delta_A = \mu_1 - \mu_2$ but also $n$.
Generally, as $n$ increases, $\beta$ decreases.

The power of a test is given by $\pi = 1 - \beta$.
We are interested in obtaining a power of $\pi = 0.8$ given all of the other
specifications by increasing sample size $n$ per population (so, the sample size
is $2 n$).

We may quickly compute an approximation of $n$ with
$$
  n = 2(z_{\alpha/2} + z_{\beta})\sigma^2 / \delta_A^2 = 123.0704,
$$
but of course we do not know $\sigma$.
So, instead, we compute $n$ with:
```{r}
sd = .7
alpha = .05
h = .8  # power
v = .25 # alternative, a practical difference
power.t.test(n=NULL,delta=v,sd=sd,sig.level=alpha,power=h,type="two.sample")
```

We see that $n' = 124.0381$.
If the power $h$ is a lower-bound, then
let $n=\lceil n' \rceil = 125$.
If the power specification is not a lower-bound, but an approximate specification,
it seems appropriate to set $n=124$.

## Part (b)
\fbox{\begin{minipage}{\textwidth}
Graph the power curve for the chosen sample size.
Explain how the power curve displays the desired properties of the test.
\end{minipage}}

```{r}
# power.curve
#
# create the power curve for the chosen sample size.
#
# arguments:
#   n: sample size,
#   sd: standard deviation
#   alpha: significance level
#   h: power (shows as a horizontal line)
#   v: specific alternative (shows as a vertical line)
#
# output:
#   graph of power curve
power.curve = function(n, sd, alpha, h, v)
{
  df = 2*(n-1)
  delta = seq(from=0,to=5*sd/sqrt(n/2),length.out = 1000)
  power = 1 - pt(qt(1-alpha/2,df),df,ncp = sqrt(n/2)*(delta/sd))
  
  plot(delta,power,type = "l",lwd=2,col="blue")
  abline(h=h,col="red",lwd=2)  
  abline(v=v,col="green",lwd=2)
}

power.curve(n=124,sd=sd,alpha=alpha,h=h,v=v)
```

First, we see that $\delta_A = `r v`$ obtains a power of $\pi(\delta_A) = 0.8$,
as specified.

Second, and this is not explicitly shown on the graph,
whether $H_0 : \delta = 0$ is true or $H_A : \delta = \delta_A$ is true, there
is a low probability of committing an error
since the $95\%$ confidence interval under the null model, approximately
$$
  \pm 1.96 \sigma / \sqrt{n} = [`r round(-1.96*sd/sqrt(124),digits=3)`,`r round(1.96*sd/sqrt(124),digits=3)`],
$$
does not intersect with an approximation of the $95\%$ confidence interval under
the alternative model,
$$
  \delta_A \pm 1.96 \sigma / \sqrt{n} = [`r round(v - 1.96*sd/sqrt(124),digits=3)`,`r round(v + 1.96*sd/sqrt(124),digits=3)`].
$$

The small region between these confidence intervals may be classified
as the "don't care" region.

Note that we do not know $\sigma$ so these CIs are approximate (and slightly
more narrow in length).
However, since $n$ is pretty large, it should be reasonably accurate, i.e.,
the limiting distribution of the $t$ distribution is the standard normal as
$n \to \infty$.

## Part (c)
\fbox{\begin{minipage}{\textwidth}
Provide a general explanation of how $\delta_A$ can be determined.
\end{minipage}}

The specific alternative $\delta_A$ is chosen to represent an effect size
that is expected (e.g., from past experience or related data), important
(difference is non-negligible), and/or practical (cost considerations).

# Problem 4
Refer back to the tensile strength example of problem 2.
Use the data from this study to perform a power analysis for a main study.
The experimenters desire a level $\alpha = .05$ test with power $\pi = .8$.

## Part (a)
\fbox{\begin{minipage}{\textwidth}
Determine the sample size for each group based on specifying the maximum
difference in means.
\end{minipage}}

```{r}
library("readxl")
library("multcomp")
data = read_excel("./handout2data.xlsx")
strength = na.omit(data$strength)
percent = na.omit(as.factor(data$percent))

pwr = .8
alpha = .05
a = 5
means = by(strength,percent,mean)
max.D = max(means) - min(means)
#summary(aov(strength~percent))
s2 = 8.06
power.anova.test(
  groups=a,
  between.var=max.D^2/2/(a-1),
  within.var=s2,
  power=pwr,
  sig.level=alpha,
  n=NULL)
```

We see that $n' = 2.534$.
We propose using a sample size of $n=3$ for each group.

## Part (b)
\fbox{\begin{minipage}{\textwidth}
Use a simulation to compute power at $n=3$ using the pilot study to specify
the model parameters.
\end{minipage}}

```{r}
sim.size = 100000
decide.Ha = rep(NA,sim.size)
n = 3

for (k in 1:sim.size)
{
  y1 = rnorm(n,means[1],sqrt(s2))
  y2 = rnorm(n,means[2],sqrt(s2))
  y3 = rnorm(n,means[3],sqrt(s2))
  y4 = rnorm(n,means[4],sqrt(s2))
  y5 = rnorm(n,means[5],sqrt(s2))
  ybar1 = mean(y1)
  ybar2 = mean(y2)
  ybar3 = mean(y3)
  ybar4 = mean(y4)
  ybar5 = mean(y5)
  var1 = var(y1)
  var2 = var(y2)
  var3 = var(y3)
  var4 = var(y4)
  var5 = var(y5)
  F.stat = n*var(c(ybar1,ybar2,ybar3,ybar4,ybar5)) / mean(c(var1,var2,var3,var4,var5))
  decide.Ha[k] = (F.stat > qf(1-alpha,a-1,a*(n-1)))
}
power = mean(decide.Ha)
power
```


## Part (c)
\fbox{\begin{minipage}{\textwidth}
Comment on the use of pilot study data in a power analysis.
\end{minipage}}

Specifying parameter values for a power analysis based on estimates from a pilot
study, without accounting for estimation error, may lead to a hypothesis test
that does not have adequate power.