---
title: 'STAT 581 - Exam 2: Due Dec 14, 2021'
author: "Alex Towell (atowell@siue.edu)"
output:
  pdf_document:
    #df_print: kable
    latex_engine: xelatex
    #keep_tex: true
    highlight: tango
header-includes:
 - \usepackage{custom}
editor_options:
  markdown:
    wrap: 80
---


```{r,echo=F}
mixed.test = function(A,B,y)
{
  av=anova(lm(y~A*B))
  F.a = av$`Mean Sq`[1]/av$`Mean Sq`[3]
  p.value = pf(F.a,df1=av$Df[1],df2=av$Df[3],lower.tail = FALSE)
  table1 = matrix(c(av$`Sum Sq`[1],av$`Sum Sq`[2],av$`Sum Sq`[3],av$`Sum Sq`[4],
                    av$Df[1],av$Df[2],av$Df[3],av$Df[4],
                    av$`Mean Sq`[1],av$`Mean Sq`[2],av$`Mean Sq`[3],av$`Mean Sq`[4]),nrow = 4)
  dimnames(table1) = list(c("Fixed Effect A","Random Effect B","Interaction AB","Error"),
                          c("SS","df","MS"))
  print(table1)
  
  table2 = matrix(c(F.a,p.value),nrow = 1)
  dimnames(table2) = list(c(""),c("F-test for fixed effect","p-value"))
  print(table2)
  
  a=nlevels(A)
  b=nlevels(B)
  n=length(y) / a / b
  
  var.hat = av$`Mean Sq`[4]
  var.interaction.hat = (av$`Mean Sq`[3]-av$`Mean Sq`[4])/n
  var.block = (av$`Mean Sq`[2]-av$`Mean Sq`[3])/n/a
  
  table3 = matrix(c(var.hat,var.interaction.hat,var.block),nrow=1)
  dimnames(table3) = list(c(""),c("error.var","interaction.var","block.var"))
  print(table3)
}
```

# Problem 1
> A randomized complete block design is used to study the effect of caliper on the measured diameters
(in mm) of ball bearings. A sample of $b=10$ ball bearings is randomly selected, and each of $a=3$ calipers
produces a measurement on each of the selected ball bearings. The data is provided as an attachment.

## P1: (c)
> Test for systematic differences in the measurements provided by the calipers.
Compute the $F_0$ statistic, and the $p$-value.
Provide an interpretation, stated in the context of the problem.

```{r, warning=F, message=F}
library("readxl")
#library("lme4")
library("lmerTest")

data = read_excel("exam2data.xlsx")
A = as.factor(na.omit(data$caliper)) # fixed effect
B = as.factor(na.omit(data$ball.bearing)) # random effect
y = na.omit(data$diameter) # response
head(data.frame(caliper=A,ball.bearing=B,diameter=y))

# fixed effect tau1 + tau2 + tau3 = 0 (for calipers)
contrasts(A)=contr.sum

random.mod = lmer(y ~ (1|B) + A)

# the anova command is used to compute the test for fixed effects.
anova(random.mod)
```

## P1: (d)
> Compute estimates of the variance components. Explain when a block design is better than a completely
randomized design.

```{r}
print(VarCorr(random.mod),comp="Variance")
```

## P1: (e)
> Compute estimates of the fixed effect parameters. Explain why block effects are modeled differently
than treatment effects in this design. Explain what treatment effect is estimable in this design.

```{r}
est = coef(summary(random.mod))[1:3,1]
est.tau.hat = c(est,0-sum(est[2:3]))
names(est.tau.hat) = c("mu","tau1","tau2","tau3")
round(est.tau.hat,digits=3)
```

# Problem 2
> Now, a mixed effects design is used to study the effect of caliper (fixed effect,
factor $A$) on the measured diameters of ball bearings. There are $a=2$ calipers
under investigation. A random sample of $b=8$ ball bearings is selected (random
effect, factor $B$), and each caliper produces $n=3$ measurements on each of the
selected ball bearings. The data is provided as an attachment.

## P2: (b)
> Create an interaction plot to display the caliper effect on measured diameter. Use a mixed model
likelihood approach to test for a systematic difference in the measurements of the two calipers.
Compute $F_0$ and the $p$-value.

```{r}
A = as.factor(na.omit(data$device)) # fixed effect
B = as.factor(na.omit(data$ball)) # random effect
y = na.omit(data$measurement) # response
contrasts(A)=contr.sum
head(data.frame(caliper=A,ball=B,measurement=y))

interaction.plot(B,A,y)
mixed.mod = lmer(y ~ A + (1|B) + (1|A:B))
anova(mixed.mod)
```

## P2: (d) and (e)
> Perform a test for caliper effects. Compute $F_A$ and the $p$-value.
> Compute the unbiased estimates of the random effect parameters.

```{r}
mixed.test(A,B,y)
```

# Problem 3
> A nested design is used to study the number of cases produced from three
bottling machines (factor $A$, fixed effect).
Four operators are randomly selected for each of the machines (nested factor $B$, random effect).
Each operator makes $n=2$ experimental runs. The data is provided as an attachment.

## P3: (d)
> Test for differences between bottling machines. Write the test statistic
$F_A$ statistic as a ratio of mean squares.

```{r,echo=F}
nested.test = function(A,B,y)
{
  av=anova(lm(y~A/B))
  ss.A = av$`Sum Sq`[1]
  ss.B = av$`Sum Sq`[2]
  ss.error = av$`Sum Sq`[3]
  df.A = av$Df[1]
  df.B = av$Df[2]
  df.error = av$Df[3]
  ms.A = ss.A / df.A
  ms.B = ss.B / df.B
  ms.error = ss.error / df.error
  F.a = ms.A / ms.B
  p.value = pf(F.a,df1=df.A,df2=df.B,lower.tail = FALSE)
  table1 = matrix(c(ss.A,ss.B,ss.error,
                    df.A,df.B,df.error,
                    ms.A,ms.B,ms.error),nrow = 3)
  dimnames(table1) = list(c("Fixed Effect A","Random Effect B(A)","Error"),
                          c("SS","df","MS"))
  print(table1)
  
  table2 = matrix(c(F.a,p.value),nrow = 1)
  dimnames(table2) = list(c(""),c("F-test for fixed effect","p-value"))
  print(table2)
  
  a=nlevels(A)
  b=nlevels(B)
  n=length(y) / a / b
  
  var.hat = ms.error
  var.B.hat = (ms.B - ms.error) / n
  
  table3 = matrix(c(var.hat,var.B.hat),nrow=1)
  dimnames(table3) = list(c(""),c("error.var","B.var"))
  print(table3)
}
```

```{r}
A = as.factor(na.omit(data$machine))
B = as.factor(na.omit(data$operator))
y = na.omit(data$cases)
nested.test(A,B,y)
```

## P3: (e)
> Explain why $\ms{E}$ is the incorrect error term to use when the nested
factor is random. In particular, comment on the pertinent sample size.

# Problem 4
> A company wishes to study the effect of promotion type $(1,2,3)$ on the sales of its crackers.
A sample of $N=15$ grocery stores is selected.
Response variable $y$ is the number of cases sold during the promotion period.
Factor $A$ is the promotion type.
Covariate $x$ is the same store sales prior to the promotion.
The data is provided as an attachment.

## P4: (a)
> Compute the estimated regression of presales on cases sold for each promotion type.
Create a scatterplot of presales versus sales for each promotion type, including
the estimated regression lines.

```{r}
prom = as.factor(na.omit(data$promotion)) # factor A
presales = na.omit(data$presales)         # predictor x
sales = na.omit(data$sales)               # response

ancova.mod = lm(sales ~ presales + prom)
est = coef(ancova.mod)
b1 = est[2]
b0 = c(est[1],est[1]+est[3],est[1]+est[4])

reg.funcs = matrix(c(b0,rep(b1,3)),nrow = 3)
dimnames(reg.funcs)=list(c("prom 1","prom 2","prom 3"),c("intercept","slope")) 
reg.funcs

# scatterplot
plot(presales[prom==1],sales[prom==1],xlab='presales',ylab='sales', pch=15,
     xlim=c(min(presales),max(presales)),
     ylim=c(min(sales),max(sales)))
points(presales[prom==2],sales[prom==2], pch=16)
points(presales[prom==3],sales[prom==3], pch=17)

abline(b0[1],b1,lty=1)
abline(b0[2],b1,lty=2)
abline(b0[3],b1,lty=3)
```

## P4: (b)
> Test for a promotion effect. Write the `ANCOVA` $F_{A|x}$ statistic using
extra sum of squares notation.
Compute $F_{A|x}$ and the $p$-value. Provide an interpretation, stated in the
context of the problem. Note the role the covariate is playing in this analysis.

```{r}
anova(ancova.mod)
```

## P4: (c)
> Compute the sample mean sales and the sample mean presales for each promotion type.
Compute the least squares means.
Explain how the information from the covariate adjusts the determination of promotion
effect.


```{r, warning=F, message=F}
library("lsmeans")
library("car")

xbar.prom = aggregate(presales, by=list(prom), FUN=mean)
ybar.prom = aggregate(sales, by=list(prom), FUN=mean)[2]
means.table = cbind(xbar.prom,ybar.prom)
colnames(means.table) = c("prom","presales.mean","sales.mean")
means.table

# overall mean
mean(presales)

# compute least squares means for adjusted means. 
lsmeans(ancova.mod,pairwise ~ prom,adjust="none")$lsmeans
```


# Code

```{r}
mixed.test = function(A,B,y)
{
  av=anova(lm(y~A*B))
  F.a = av$`Mean Sq`[1]/av$`Mean Sq`[3]
  p.value = pf(F.a,df1=av$Df[1],df2=av$Df[3],lower.tail = FALSE)
  table1 = matrix(c(av$`Sum Sq`[1],av$`Sum Sq`[2],av$`Sum Sq`[3],av$`Sum Sq`[4],
                    av$Df[1],av$Df[2],av$Df[3],av$Df[4],
                    av$`Mean Sq`[1],av$`Mean Sq`[2],av$`Mean Sq`[3],av$`Mean Sq`[4]),nrow = 4)
  dimnames(table1) = list(c("Fixed Effect A","Random Effect B","Interaction AB","Error"),
                          c("SS","df","MS"))
  print(table1)
  
  table2 = matrix(c(F.a,p.value),nrow = 1)
  dimnames(table2) = list(c(""),c("F-test for fixed effect","p-value"))
  print(table2)
  
  a=nlevels(A)
  b=nlevels(B)
  n=length(y) / a / b
  
  var.hat = av$`Mean Sq`[4]
  var.interaction.hat = (av$`Mean Sq`[3]-av$`Mean Sq`[4])/n
  var.block = (av$`Mean Sq`[2]-av$`Mean Sq`[3])/n/a
  
  table3 = matrix(c(var.hat,var.interaction.hat,var.block),nrow=1)
  dimnames(table3) = list(c(""),c("error.var","interaction.var","block.var"))
  print(table3)
}

nested.test = function(A,B,y)
{
  av=anova(lm(y~A/B))
  ss.A = av$`Sum Sq`[1]
  ss.B = av$`Sum Sq`[2]
  ss.error = av$`Sum Sq`[3]
  df.A = av$Df[1]
  df.B = av$Df[2]
  df.error = av$Df[3]
  ms.A = ss.A / df.A
  ms.B = ss.B / df.B
  ms.error = ss.error / df.error
  F.a = ms.A / ms.B
  p.value = pf(F.a,df1=df.A,df2=df.B,lower.tail = FALSE)
  table1 = matrix(c(ss.A,ss.B,ss.error,
                    df.A,df.B,df.error,
                    ms.A,ms.B,ms.error),nrow = 3)
  dimnames(table1) = list(c("Fixed Effect A","Random Effect B(A)","Error"),
                          c("SS","df","MS"))
  print(table1)
  
  table2 = matrix(c(F.a,p.value),nrow = 1)
  dimnames(table2) = list(c(""),c("F-test for fixed effect","p-value"))
  print(table2)
  
  a=nlevels(A)
  b=nlevels(B)
  n=length(y) / a / b
  
  var.hat = ms.error
  var.B.hat = (ms.B - ms.error) / n
  
  table3 = matrix(c(var.hat,var.B.hat),nrow=1)
  dimnames(table3) = list(c(""),c("error.var","B.var"))
  print(table3)
}
```
