---
title: 'Discrete Multivariate Analysis - 579 - Final Exam - Part 2'
author: "Alex Towell (atowell@siue.edu)"
fontfamily: mathpazo
fontsize: 11pt
geometry: margin=1in
output:
  pdf_document:
    keep_tex: true
    latex_engine: xelatex
    df_print: kable
#    number_sections: true
  html_document:
    df_print: paged
header-includes:
 - \usepackage{alex}
---

```{r, include=FALSE}
options(tinytex.engine_args = '-shell-escape')
```

Applicants for graduate school are classified according to department, sex, and
admission status.
A goal of the study is to determine the role an applicant's sex plays in the
determination of admission status.

```{r}
# department and sex are defined through indicator variables
grad.data <- data.frame(dep=c(0,0,1,1),
                        sex=c(0,1,0,1),
                        yes=c(235,38,122,103),
                        no=c(35,7,93,69))

# define row sample sizes
grad.data$n = grad.data$yes + grad.data$no
# define the interaction variable
grad.data$dep.sex = grad.data$dep * grad.data$sex

print(grad.data)
```

# Problem 1
\fbox{\begin{minipage}{.9\textwidth}
Define a main effects logistic regression model $M$ having two binary input
variables.
Include notation for the design matrix $X$, and the parameter vector $\beta$.
Provide an interpretation for each of the effect parameters in $\beta$, stated
in the context of the problem.
\end{minipage}}

The data is cross-sectional data given by
$$
    (\v{x_1},y_1),(\v{x_2},y_2),\ldots,(\v{x_n},y_n)
$$
where $\v{x_i} = (1,x_{i 1},x_{i 2})'$.

The response variables $\{y_1,\ldots,y_n\}$ are given by
$$
  y =
  \begin{cases}
    1 & \text{if admit=yes}\\
    0 & \text{if admit=no}
  \end{cases}
$$

The explanatory indicator variables $x_1$ and $x_2$ are respectively given by
$$
  x_1 =
  \begin{cases}
    1 & \text{if department=non-science}\\
    0 & \text{if department=science}
  \end{cases}
$$
and
$$
  x_2 =
  \begin{cases}
    1 & \text{if sex=female}\\
    0 & \text{if sex=male}
  \end{cases}
$$

Then, $\pi$ denotes the probability of being admitted,
$$
  \pi(\v{x}) = \Pr(y = \text{yes} \;|\; \v{x}).
$$
That is, $\pi$ models probability as a function of $\v{x}$.

The probability model is given by
\[
  y_i \sim \operatorname{BIN}(1,\pi(\v{x_i}))
\]

The main effects model $M$ is given by
$$
  \operatorname{logit}(\pi(\v{x})) = \v{x}' \v{\beta},
$$
where the parameter vector \(\v{\beta}\) of dimension \(3 \times 1\) is
given by
\[
\v{\beta} =
\begin{pmatrix}
  \beta_0\\
  \beta_1\\
  \beta_2
\end{pmatrix}
\].

The parameter $\beta_1$ is the partial effect of department
(non-science - science) on admission and the parameter $\beta_2$ is the partial
effect of sex (female - male) on admission.

In the main effects model $M$, we assume the effect of $x_1$ is the same at
both levels of $x_2$ and vice versa, i.e.,
$$
  \theta(\text{science}:\text{non-science} \;|\; \text{male}) = \theta(\text{science}:\text{non-science} \; | \;\text{female}) = \exp(\beta_1)
$$
and
$$
  \theta(\text{male}:\text{female} \;|\; \text{science}) = \theta(\text{male}:\text{female} \; | \; \text{non-science}) = \exp(\beta_2).
$$

The logistic regression model for $M$ is given by
\[
  \operatorname{logit}(\pi(\v{x})) = \v{x_i}' \v{\beta} = \beta_0 + \beta_1 x_{i 1} + \beta_2 x_{i 2},
\]
such that if we solve for \(\pi(\v{x_i})\) we get the result
\[
  \pi(\v{x_i}) = \frac{\exp(\v{x_i}' \v{\beta})}{1+\exp(\v{x_i}' \v{\beta})}.
\]

The design matrix is given by
$$
\mat{X}
=
\begin{pmatrix}
  1 & 0 & 0\\
  1 & 0 & 1\\
  1 & 1 & 0\\
  1 & 1 & 1
\end{pmatrix}.
$$

Since we are dealing with Boolean input variables $\v{x} = (1,x_1,x_2)$,
we only need to count the number of times each discrete input in the design
matrix occurs, say $n_j$ for the $j$-th row, denoted by $\v{x_j}$, and let the
corresponding $y_j$ denote the number of times the response was \emph{yes} for
$\v{x_j}$.

Then, the probabilistic model for $y_j$ is the binomial distribution,
$$
  y_j \sim \operatorname{BIN}(n_j, \pi(\v{x_j}))
$$
and we estimate $\v{\beta}$ by maximizing
$$
  \hat{\v{\beta}} = \argmax_{\v{\beta}} \prod_{i=1}^4 \pi_i(\v{\beta})^{y_i}(1-\pi_i(\v{\beta}))^{n_i - y_i}.
$$

# Problem 2
\fbox{\begin{minipage}{.9\textwidth}
Provide notation for the design matrix $X_S$ and parameter vector $\beta_S$
for the saturated model $M_S$.
Provide a brief description of an interaction effect.
\end{minipage}}

For the saturated model $M_S$ (interaction model), we must include all possible
effects and interactions, which means relative to the main effects model $M$ we
add account for interaction effects between $x_1$ and $x_2$, i.e.,
$$
  \operatorname{logit}(\pi(\v{x})) = \v{x}' \v{\beta_S} = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \beta_3 x_1 x_2
$$
where the parameter vector \(\v{\beta_S}\) of dimension \(4 \times 1\) is
given by
\[
\v{\beta_S} =
\begin{pmatrix}
  \beta_0\\
  \beta_1\\
  \beta_2\\
  \beta_3
\end{pmatrix}
\].

The interaction effect between department and sex ($x_1$ and $x_2$) occurs when
the effect of an input depends on the level of the input of the other input.
In the main effects model $M$, we assume the effect of $x_1$ is the same at
both levels of $x_2$ and vice versa, i.e.,
\begin{align*}
  \theta(\text{science}:\text{non-science} \;|\; \text{male})    &= e^{\beta_1},\\
  \theta(\text{science}:\text{non-science} \; | \;\text{female}) &= e^{\beta_1+\beta_3},\\
  \theta(\text{male}:\text{female} \;|\; \text{science})         &= e^{\beta_2},\\
  \theta(\text{male}:\text{female} \; | \; \text{non-science})   &= e^{\beta_2+\beta_3}.
\end{align*}

The design matrix for $M_S$ is given by
$$
\mat{X_S} = 
\begin{pmatrix}
  1 & 0 & 0 & 0\\
  1 & 0 & 1 & 0\\
  1 & 1 & 0 & 0\\
  1 & 1 & 1 & 1
\end{pmatrix}.
$$


<!-- According to the design matrix $\mat{X_S}$, the parameter $\beta_3$ is the -->
<!-- mixed effect that department (non-science or science) and sex (female or male) -->
<!-- have on admission, but the mixed effect can only distinguish between two -->
<!-- cases: -->
<!-- \begin{enumerate} -->
<!-- \item Both $x_1$ and $x_2$ set to $1$. -->
<!-- \item Both of them are not set to $1$ (at least one of them is set to $0$). -->
<!-- \end{enumerate} -->

# Problem 3
\fbox{\begin{minipage}{.9\textwidth}
For each of the models $M_O$, $M_1$, $M_2$, provide notation for the design
matrix and a brief description of the model effects, stated in the context of
the problem.
\end{minipage}}

The design matrix for $M_O$, the independence model, is given by
$$
\mat{X_O} = 
\begin{pmatrix}
  1\\
  1\\
  1\\
  1
\end{pmatrix}.
$$
This model assumes that neither department nor sex has an effect on admission,
i.e., $\operatorname{logit}(\pi) = \beta_0$.

The design matrix for $M_1$ is given by
$$
\mat{X_1} = 
\begin{pmatrix}
  1 & 0\\
  1 & 0\\
  1 & 1\\
  1 & 1
\end{pmatrix}.
$$
This model considers only the marginal effect of department on admission,
i.e., $\operatorname{logit}(\pi(x_1)) = \beta_0 + \beta_1 x_1$.

The design matrix for $M_2$ is given by
$$
\mat{X_2} = 
\begin{pmatrix}
  1 & 0\\
  1 & 1\\
  1 & 0\\
  1 & 1
\end{pmatrix}.
$$
This model considers only the marginal effect of sex on admission,
i.e., $\operatorname{logit}(\pi(x_1)) = \beta_0 + \beta_2 x_2$.

# Problem 4
\fbox{\begin{minipage}{.9\textwidth}
Compute the deviance statistic $D$, and give degrees of freedom $\Delta df$, for
each of the models $M_O,M_1,M_2,M,M_S$ from the grad school data.
Provide a general form for the statistic $G^2$, and the degrees of freedom for
the reference chi-square distribution, for testing a reduced model $M_R$ against
a full model $M_F$.
\end{minipage}}

$G^2(M\,|\,M_S) = D(M)$ is called the deviance for testing model $M$
goodness-of-fit.

The likelihood ratio statistic $G^2$ for testing a reduced model $M_R$ against
a full model $M_F$ is given by
\begin{align*}
  G^2(M_R\,|\,M_S)
    &= [-2 L_R]-[-2 L_F]\\
    &= \left([-2 L_R]-[-2 L_S]\right)-\left([-2 L_F]-[-2 L_S] \right)\\
    &= G^2(M_R\,|\,M_S)-G^2(M_F\,|\,M_S)\\
    &= D(M_R) - D(M_F).
\end{align*}

The degree of freedom is given by
$$
  df = P_F - P_R = (P_S - P_R) - (P_S - P_F) = \Delta df_R - \Delta df_F,
$$
and so the reference distribution is $\chi^2$ with $df = \Delta df_R - \Delta df_F$.

```{r}
# fit each of the candidate models
m.s = glm(yes/n ~ dep+sex+sex*dep,weights=n,family=binomial,data=grad.data)
m.12 = glm(yes/n ~ dep+sex,weights=n,family=binomial,data=grad.data)
m.1 = glm(yes/n ~ dep,weights=n,family=binomial,data=grad.data)
m.2 = glm(yes/n ~ sex,weights=n,family=binomial,data=grad.data)
m.0 = glm(yes/n ~ 1,weights=n,family=binomial,data=grad.data)

# create a table for candidate model deviances
# the rows represent the model, the deviance, and the degrees of freedom
deviance.table=matrix(c(
  0,m.12$deviance,m.1$deviance,m.2$deviance,m.0$deviance,
  0,m.12$df.residual,m.1$df.residual,m.2$df.residual,m.0$df.residual),
  nrow=5)
dimnames(deviance.table) = list(c("MS","M","M1","M2","MO"),c("deviance","delta.df"))
print(deviance.table)
```

# Problem 5
\fbox{\begin{minipage}{.9\textwidth}
Compute the likelihood statistic $G^2$ for testing reduced model $M$ against
full model $M_S$ from the grad school data, and provide an interpretation in the
context of the problem.
\end{minipage}}

```{r}
# test the main effects model against the interaction (saturated) model
anova(m.12,m.s,test="LRT")
```

The statistic $G^2(M\,|\,M_S) = D(M) = 0.459$, which has a $p$-value of $0.498$.
Thus, we conlcude the main effects model (no interaction on inputs) is
compatible with the observed data.

# Problem 6
\fbox{\begin{minipage}{.9\textwidth}
Compute the likelihood statistic $G^2$ for testing reduced model $M_O$ against
full model $M_2$ from the grad school data, and provide an interpretation in the
context of the problem.
\end{minipage}}

We are testing for a marginal effect of $x_2$ (sex), ignoring the effects of
$x_1$ (department).

```{r}
# test for the input 2 effect using a marginal effect test
anova(m.0,m.2,test="LRT")
```

The statistic $G^2(M_0\,|\,M_2) = 5.317$ with $df=1$ has a $p$-value around
$0.021$.
Thus, we conlcude that the data supports the model where sex has a marginal
effect on admission.

# Problem 7
\fbox{\begin{minipage}{.9\textwidth}
Compute the likelihood statistic $G^2$ for testing reduced model $M_1$ against
full model $M$ from the grad school data, and provide an interpretation in the
context of the problem.
Include an explanation of how this test differs from that of the previous
problem.
\end{minipage}}

We are testing for a partial effect of $x_2$ (sex), adjusting for the effect of
$x_1$ (department).

```{r}
anova(m.1,m.12,test="LRT")
```

The statistic $G^2(M_1\,|\,M) = 67.420$ with $df=1$ has $p\text{-value} = 0.704$,
thus we conclude that when we adjust for $x_1$ (department), the data supports
the model where $x_2$ (sex) does not have a significant effect.

How are we to interpret the results?
We have shown that sex has a marginal effect on admission, but its effect when
adjusting for department is neglibible.

We make the following observation.
Males are more likely to choose science, and science department is more likely
to accept.
Females are more likely to choose non-science, and non-science department is
less likely to accept.
Thus, we see that overall, females are less likely to be admitted, thus showing
that sex has a marginal effect on admission.
However, the probability of admission is the same for both sexes when we
adjust for department.

Thus, model $M_1$ (department) is the major effect.

# Problem 8
\fbox{\begin{minipage}{.9\textwidth}
Compute estimates of the response probabilities based on model $M_1$ from the
grad school data, and provide an interpretation in the context of the problem.
\end{minipage}}

```{r}
# compute probability estimates under model M1
pred.1 = predict(m.1,type = "response")

prob.table = matrix(c(pred.1),nrow = 4)
dimnames(prob.table) = list(c("science male","science female",
                              "non-science male","non-science female"),
                            c("prob.1"))
print(prob.table,digits = 4)
```

Under model $M_1$, we estimate the probability of admission into the science
department is $0.8667$ and we estimate probability of admission into the
non-science department is $0.5814$.

Note that under this model, sex has no effect on the probability of admission.
This model is justified on the basis of the data and the previous tests
we conducted.